---
title: "HNSCC"
author: "Yufei Wang"
date: "2024-06-04"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```


```{r library used}
library(R.utils)
library(tidyverse)
library(knitr)
library(kableExtra)
library(readr)
library(data.table)
library(arrow)
library(reticulate)
library(stringr)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(ENmix)
library(parallel)
library(pheatmap)
library(webshot)
library(xml2)
library(rvest)
library(htmltools)
library(gridExtra)
library(grid)
library(writexl)
library(survival)
library(survminer)
library(magick)
#library(minfi)
library(wateRmelon)
#library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
```


```{python Create the Parquet file for methylation data}
# import pandas as pd
# import os
# import logging
# import pyarrow 
# import fastparquet
# 
# # Configure logging
# logging.basicConfig(level=logging.INFO, format='%(asctime)s - %(levelname)s - %(message)s')
# 
# # Define the file names
# txt_file_name = '/Users/yufeiwang/Desktop/HNSCC/methylation_level3_within_bioassay.txt'
# parquet_file_name = '/Users/yufeiwang/Desktop/HNSCC/methylation_level3_within_bioassay.parquet'
# 
# # Function to process and save chunks individually
# def process_and_save_chunk(chunk, parquet_file, append=False):
#     # Convert all columns to string type to avoid type conversion issues
#     chunk = chunk.astype(str)
#     
#     # Write to Parquet file
#     if append:
#         chunk.to_parquet(parquet_file, engine='fastparquet', append=True)
#     else:
#         chunk.to_parquet(parquet_file, engine='fastparquet')
# 
# # Function to load the data in chunks and save as Parquet
# def load_and_save_as_parquet(txt_file, parquet_file, chunk_size=100000):
#     logging.info("Starting to load the dataset in chunks")
# 
#     try:
#         first_chunk = True
#         for chunk in pd.read_csv(txt_file, delimiter='\t', chunksize=chunk_size, low_memory=False):
#             logging.info(f"Processed chunk with {len(chunk)} rows")
#             process_and_save_chunk(chunk, parquet_file, append=not first_chunk)
#             first_chunk = False
#         
#         logging.info("Dataset loaded and saved successfully")
#     except Exception as e:
#         logging.error(f"Error loading dataset: {e}")
#         raise
```


```{r load data}
# Load data for clinical, methylation, and mRNA 
clinical_all_CDEs <- fread("clinical_All_CDEs.txt", sep = "\t", header = TRUE)

# Open the Parquet file and read it into a dataframe (methylation data)
methyl_level3 <- read_parquet("methylation_level3_within_bioassay.parquet")
```


```{r preprocess the clinical data}
# Transpose the clinical data
clinical_transposed <- t(clinical_all_CDEs)

clinical_transposed <- as.data.frame(clinical_transposed)

colnames(clinical_transposed) <- clinical_transposed[1,]

clinical_transposed <- clinical_transposed[-1,]

# Select necessary variables
clinical_focused <- clinical_transposed[, c(9, 12, 21:23, 46, 48, 54, 114)]

# Save the cleaned dataframe to a new Parquet file
write_parquet(clinical_focused, "clinical_focused.parquet")

# Filter data
clinical_focused_filtered <- clinical_focused %>%
  filter(!is.na(clinical_m), !is.na(clinical_n), !is.na(clinical_stage), hpv_status != "indeterminate", tumor_tissue_site == "head and neck")
```


```{r export .csv file for high school students}
clinical_export <- clinical_focused[, c(1, 2, 6, 8, 9)]

write.csv(clinical_export, file = "/Users/yufeiwang/Desktop/HNSCC/clinical_data.csv", row.names = FALSE)
```



## Overview of Clinical Data


```{r Table of anatomic neoplasm subdivision and HPV status}
subdivision <- as.factor(clinical_focused_filtered$anatomic_neoplasm_subdivision)

is_hpv <- as.factor(clinical_focused_filtered$hpv_status)

table1 <- table(subdivision, is_hpv)

table1_margin <- addmargins(table1)

kable(table1_margin, format = "html",
      caption = "Cross-Tabulation of Anatomic Neoplasm Subdivision and HPV Status") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                    full_width = FALSE)
```


```{r Table of clinical_m and HPV status}
# Distant metastasis (M)
# MX: Metastasis cannot be measured.
# M0: Cancer has not spread to other parts of the body.
# M1: Cancer has spread to other parts of the body.

clinical_m <- as.factor(clinical_focused_filtered$clinical_m)

table2 <- table(clinical_m, is_hpv)

table2_margin <- addmargins(table2)

kable(table2_margin, format = "html", 
      caption = "Cross-Tabulation of clinical_m and HPV Status") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                    full_width = FALSE)
```


```{r Table of clinical_n and HPV status}
# Regional lymph nodes (N)
# NX: Cancer in nearby lymph nodes cannot be measured.
# N0: There is no cancer in nearby lymph nodes.
# N1, N2, N3: Refers to the number and location of lymph nodes that contain cancer. The higher the number after the N, the more lymph nodes that contain cancer.

clinical_n <- as.factor(clinical_focused_filtered$clinical_n)

table3 <- table(clinical_n, is_hpv)

table3_margin <- addmargins(table3)

kable(table3_margin, format = "html", 
      caption = "Cross-Tabulation of clinical_n and HPV Status") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                    full_width = FALSE)
```


```{r Table of clinical_stage and HPV status}
clinical_stg <- as.factor(clinical_focused_filtered$clinical_stage)

table4 <- table(clinical_stg, is_hpv)

table4_margin <- addmargins(table4)

kable(table4_margin, format = "html", 
      caption = "Cross-Tabulation of clinical_stage and HPV Status") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                    full_width = FALSE)
```


```{r Table of gender and HPV status}
gender <- as.factor(clinical_focused_filtered$gender)

table5 <- table(gender, is_hpv)

table5_margin <- addmargins(table5)

kable(table5_margin, format = "html", 
      caption = "Cross-Tabulation of gender and HPV Status") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                    full_width = FALSE)
```


```{r Table of histological_type and HPV status}
histological_type <- as.factor(clinical_focused_filtered$histological_type)

table6 <- table(histological_type, is_hpv)

table6_margin <- addmargins(table6)

kable(table6_margin, format = "html", 
      caption = "Cross-Tabulation of histological_type and HPV Status") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                    full_width = FALSE)
```


```{r Preprocess the methylation data from Parquet file}
# Open the Parquet file
dataset <- open_dataset('/Users/yufeiwang/Desktop/HNSCC/methylation_level3_within_bioassay.parquet')

# Read the first few rows
first_few_rows <- as.data.frame(dataset %>% slice_head(n = 10))

# Identify the columns of interest
initial_columns <- colnames(first_few_rows)[c(1, 3, 4)]

# Identify Beta_value columns
beta_value_columns <- names(first_few_rows)[which(first_few_rows[1, ] == "Beta_value")]

# Combine the columns to keep
columns_to_keep <- c(initial_columns, beta_value_columns)

# Filter the dataset to keep only the relevant columns
filtered_dataset <- dataset %>%
  dplyr::select(all_of(columns_to_keep)) 

# Convert to data frame
cleaned_data <- as.data.frame(filtered_dataset)

# Rename the second to third columns to have blank headers
colnames(cleaned_data)[2:3] <- c("Gene_Symbol", "Chromosome")

# Save the cleaned dataframe to a new Parquet file
write_parquet(cleaned_data, "cleaned_methylation_data.parquet")
```


## Overview of Methylation Data


Number of patients need to be the same before merging methylation data and the clinical data, thus I check the duplicate patient ID in the methylation data and find that there are 52 of the duplicated patient ID. Then I check those duplicate data to ensure the sample type is solid tumor ("01"), the data is cleaned if sample type is not solid tumor, if both are solid tumor with the same beta value, only one column is kept, if both are solid tumor but with different beta value, we will calculated the average and store it in one column.


### Step 1: Check duplicate patient ID


```{r check duplicates in the methylation data}
extract_id <- function(header) {
  str_sub(header, 9, 12)
}

# Extract column headers starting from the 4th column
headers <- colnames(cleaned_data)[4:ncol(cleaned_data)]

# Extract the specific substring from each header
extracted_ids <- sapply(headers, extract_id)

# Check for duplicates
duplicate_ids <- extracted_ids[duplicated(extracted_ids)]

# Count the number of unique and duplicate IDs
unique_ids <- unique(extracted_ids)
num_unique_ids <- length(unique_ids)
num_duplicate_ids <- length(duplicate_ids)

# Print the results
cat("Number of unique IDs:", num_unique_ids, "\n")
cat("Number of duplicate IDs:", num_duplicate_ids, "\n")
```


### Step 2: Remove data that sample type is not solid tumor ("01") and check the duplicate again


```{r ensure all the sample type is solid tumor ("01")}
# Function to extract the 14th and 15th digits from the column headers
extract_digits_14_15 <- function(header) {
  str_sub(header, 14, 15)
}

# Extract column headers starting from the 4th column
headers <- colnames(cleaned_data)[4:ncol(cleaned_data)]

# Extract the 14th and 15th digits from each header
digits_14_15 <- sapply(headers, extract_digits_14_15)

# Filter out columns where the 14th and 15th digits are not "01"
valid_columns <- headers[digits_14_15 == "01"]
removed_columns <- headers[digits_14_15 != "01"]

cleaned_data <- cleaned_data %>% select(all_of(c(colnames(cleaned_data)[1:3], valid_columns)))

# Print the removed columns
cat("Removed columns (sample type is not '01'):\n")
print(removed_columns)

# Extract the specific substring from each header
headers <- colnames(cleaned_data)[4:ncol(cleaned_data)]
extracted_ids <- sapply(headers, extract_id)

# Find duplicate IDs
duplicate_ids <- extracted_ids[duplicated(extracted_ids)]

# Print the results
cat("Number of unique IDs:", length(unique(extracted_ids)), "\n")
cat("Number of duplicate IDs:", length(duplicate_ids), "\n")

# If there are duplicates, print them
if (length(duplicate_ids) > 0) {
  cat("Duplicate IDs:\n")
  print(duplicate_ids)
}
```


### Step 2: Check sample type again


```{r check sample type again}
remaining_headers <- colnames(cleaned_data)[4:ncol(cleaned_data)]
remaining_digits_14_15 <- sapply(remaining_headers, extract_digits_14_15)

if (all(remaining_digits_14_15 == "01")) {
  cat("All remaining columns have '01' as the sample type.\n")
} else {
  cat("Some remaining columns do not have '01' as the sample type:\n")
  print(remaining_headers[remaining_digits_14_15 != "01"])
}

# Save the cleaned dataframe to a new parquet file
write_parquet(cleaned_data, "cleaned_methylation_data.parquet")
```


## Merge Methylation Data and Clinical Data


### Check patient ID correspondance


```{python remove the first row for cleaned methylation data}
# import pandas as pd
# import pyarrow.parquet as pq
# 
# # Define the path to your Parquet file
# parquet_file = "/Users/yufeiwang/Desktop/HNSCC/cleaned_methylation_data.parquet"
# output_file = "/Users/yufeiwang/Desktop/HNSCC/cleaned_methylation_data_no_first_row.parquet"
# 
# # Read the Parquet file into a DataFrame
# df = pd.read_parquet(parquet_file)
# 
# # Remove the first row
# df_no_first_row = df.iloc[1:].reset_index(drop=True)
# 
# # Save the modified DataFrame to a new Parquet file
# df_no_first_row.to_parquet(output_file)
# 
# print("First row removed and new Parquet file saved.")
```


```{r preparation for methylation and clinical data to merge together}
# transpose the clinical data
clinical_focused_merge <- t(clinical_focused)

clinical_focused_merge <- as.data.frame(clinical_focused_merge)

cleaned_data <- open_dataset("cleaned_methylation_data_no_first_row.parquet")

cleaned_data <- as.data.frame(cleaned_data)

# Rename headers for each column starting from the 4th
rename_header <- function(header) {
  tolower(str_sub(header, 1, 12))
}

new_headers <- sapply(colnames(cleaned_data)[4:ncol(cleaned_data)], rename_header)

colnames(cleaned_data)[4:ncol(cleaned_data)] <- new_headers

# Save the new dataframe to cleaned_data_merge
cleaned_data_merge <- cleaned_data

# Save the cleaned dataframe to a new parquet file
write_parquet(cleaned_data_merge, "cleaned_methylation_data_merge.parquet")
```


```{python calculate range and summation}
# import pandas as pd
# import pyarrow as pa
# import pyarrow.parquet as pq
# from tqdm import tqdm
# 
# # Define the path to the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/cleaned_methylation_data_merge.parquet"
# output_file = "/Users/yufeiwang/Desktop/HNSCC/cleaned_methylation_data_final.parquet"
# 
# # Read the Parquet file into a DataFrame
# df = pd.read_parquet(input_file)
# 
# # Replace "nan" with empty strings in the patient columns
# patient_columns = df.columns[3:]
# df[patient_columns] = df[patient_columns].apply(lambda col: col.map(lambda x: '' if pd.isna(x) else x))
# 
# # Initialize Low and High columns with NaN values
# low_high_df = pd.DataFrame({'Low': [''] * len(df), 'High': [''] * len(df)})
# df = pd.concat([df.iloc[:, :3], low_high_df, df.iloc[:, 3:]], axis=1)
# 
# # Calculate Low and High values for each row, using tqdm for progress tracking
# for index, row in tqdm(df.iterrows(), total=df.shape[0], desc="Processing rows"):
#     values = pd.to_numeric(row[5:], errors='coerce').dropna()
#     if not values.empty:
#         df.at[index, 'Low'] = values.min()
#         df.at[index, 'High'] = values.max()
# 
# # Ensure 'Low' and 'High' columns are strings
# df['Low'] = df['Low'].apply(lambda x: '' if pd.isna(x) else str(x))
# df['High'] = df['High'].apply(lambda x: '' if pd.isna(x) else str(x))
# 
# # Add the 'Sum' row as the new first row
# sum_row = [''] * 5 + df.iloc[:, 5:].apply(pd.to_numeric, errors='coerce').sum().tolist()
# sum_df = pd.DataFrame([sum_row], columns=df.columns)
# sum_df.index = ['Sum']
# df = pd.concat([sum_df, df]).reset_index(drop=True)
# 
# # Ensure all patient columns are numeric before saving to Parquet
# df[patient_columns] = df[patient_columns].apply(pd.to_numeric, errors='coerce')
# 
# # Convert all columns to strings to avoid type conversion issues
# df = df.astype(str)
# 
# # Save the modified DataFrame to a new Parquet file
# df.to_parquet(output_file)
# 
# print("All operations completed and new Parquet file saved.")
```


```{python adjust the calculated data frame}
# import pandas as pd
# import pyarrow.parquet as pq
# from tqdm import tqdm
# 
# # Define the path to the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/cleaned_methylation_data_final.parquet"
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# methyl_merge_calculated = pq.read_table(input_file).to_pandas()
# 
# # Rename the first entry in the 'Hybridization REF' column to "Summation"
# print("Renaming the first entry in 'Hybridization REF' column to 'Summation'...")
# methyl_merge_calculated.at[0, 'Hybridization REF'] = 'Summation'
# 
# # Extract the first row, starting from the 6th column
# first_row_data = methyl_merge_calculated.iloc[0, 5:]
# 
# # Convert the first row data to numeric for calculation
# print("Converting first row data to numeric for calculation...")
# numeric_first_row_data = pd.to_numeric(first_row_data, errors='coerce')
# 
# # Find the minimum and maximum values in the first row
# print("Finding the minimum and maximum values in the first row...")
# min_value = numeric_first_row_data.min()
# max_value = numeric_first_row_data.max()
# 
# # Store the minimum and maximum values in the 'Low' and 'High' columns
# print("Storing the minimum and maximum values in 'Low' and 'High' columns...")
# methyl_merge_calculated.at[0, 'Low'] = str(min_value)
# methyl_merge_calculated.at[0, 'High'] = str(max_value)
# 
# # Save the modified DataFrame back to Parquet file
# output_file = "/Users/yufeiwang/Desktop/HNSCC/cleaned_methylation_data_final.parquet"
# print("Saving the modified DataFrame back to Parquet file...")
# methyl_merge_calculated.to_parquet(output_file, index=False)
# 
# print("All operations completed and new Parquet file saved.")

```


```{r adjust the calculated data frame and append to the clinical data}
methyl_merge_calculated <- open_dataset("cleaned_methylation_data_final.parquet")

methyl_merge_calculated <- as.data.frame(methyl_merge_calculated)

sorted_clinical_focused <- clinical_focused[order(rownames(clinical_focused)), ]

methyl_sum_row <- fread("first_row_extracted.csv")

methyl_sum_row <- t(methyl_sum_row)

colnames(methyl_sum_row) <- "Summation"

sorted_methyl_sum <- methyl_sum_row[order(rownames(methyl_sum_row)), ]

sorted_methyl_sum <- as.data.frame(sorted_methyl_sum)

colnames(sorted_methyl_sum) <- "Summation"

same_names <- all(rownames(sorted_clinical_focused) == rownames(sorted_methyl_sum))

if (same_names) {
  clinical_focused_combined <- cbind(sorted_methyl_sum, sorted_clinical_focused)
} else {
  print("The row names of the datasets are not the same or not in the same order.")
}
```


```{r box plot for combined clinical data}
# Convert the Summation column to numeric
clinical_focused_combined$Summation <- as.numeric(clinical_focused_combined$Summation)

# Check for any conversion warnings
if (any(is.na(clinical_focused_combined$Summation))) {
  warning("There are NA values in the Summation column.")
}

# Extract patient IDs where hpv_status is "indeterminate"
indeterminate_ids <- clinical_focused_combined %>% 
  filter(hpv_status == "indeterminate") %>% 
  rownames()

# Remove indeterminate value (2 rows removed)
clinical_focused_combined <- clinical_focused_combined %>% 
  filter(hpv_status != "indeterminate")

# Create the box plot
ggplot(clinical_focused_combined, aes(x = hpv_status, y = Summation)) +
  geom_boxplot() +
  labs(title = "Box Plot of Beta_value Summation by HPV Status",
       x = "HPV Status",
       y = "Beta-value Summation") +
  theme_minimal()

#write_parquet(clinical_focused_combined, "clinical_focused_combined_summation.parquet")

# Save the plot
ggsave("boxplot_summation_by_hpv_status_before_removal.png")
```


```{r t-test for hpv_positive and hpv_negative groups}
# Perform t-test between HPV positive and HPV negative groups
t_test_result <- t.test(Summation ~ hpv_status, data = clinical_focused_combined)

print(t_test_result)
```


```{r wilcoxon test before removal}
wilcox_test_before <- wilcox.test(Summation ~ hpv_status, data = clinical_focused_combined)

print(wilcox_test_before)
```


We will remove two patient's record beacuse they have "indeterminate" value for hpv status. (531 columns left, 526 patients left)

```{r remove nan rows and 2 columns (indeterminate hpv status)}
# Remove columns with IDs in indeterminate_ids
methyl_merge_calculated <- methyl_merge_calculated %>%
  select(-all_of(indeterminate_ids))

# Save the cleaned dataframe to a new parquet file
write_parquet(methyl_merge_calculated, "methyl_calculated_removed_indeter_hpv.parquet")
```


Now we will remove those rows where all beta_value is "nan" in Python


```{python}
# import pandas as pd
# from tqdm import tqdm
# 
# # Define the path to the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_calculated_removed_indeter_hpv.parquet"
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet"
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# df = pd.read_parquet(input_file)
# 
# # Ensure Low column is treated as a string to detect empty strings correctly
# df['Low'] = df['Low'].astype(str)
# 
# # Remove rows where the Low column is empty or NaN
# print("Removing rows with empty Low values...")
# df_cleaned = df[df['Low'].notna() & (df['Low'] != '')]
# 
# # Save the cleaned DataFrame to a new Parquet file
# print("Saving the cleaned DataFrame to a new Parquet file...")
# df_cleaned.to_parquet(output_file, index=False)
# 
# print("All operations completed and new Parquet file saved.")

```


```{r}
# This is the data that have been cleaned for both hpv "indeterminate" and "nan" rows
methyl_merge_calculated_cleaned <- open_dataset("methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet")
methyl_merge_calculated_cleaned <- as.data.frame(methyl_merge_calculated_cleaned)
```


Now I use Python to split a subset of the data for 15th percentile calculation.


```{python}
# import pandas as pd
# 
# # Define the path to the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet"
# output_file = "/Users/yufeiwang/Desktop/HNSCC/subdataset_for_15_perc.parquet"
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# df = pd.read_parquet(input_file)
# 
# # Extract the sub-dataset from the 2nd row and starting from the 6th column
# sub_df = df.iloc[1:, 5:]
# 
# # Save the sub-dataset to a new Parquet file
# print("Saving the sub-dataset to a new Parquet file...")
# sub_df.to_parquet(output_file, index=False)
# 
# print("All operations completed and new Parquet file saved.")
```


Load the sub-data:

```{r}
methyl_sub <- open_dataset("subdataset_for_15_perc.parquet")
methyl_sub <- as.data.frame(methyl_sub)
```


Calculate the 15th percentile among all the beta_values for all CpGs.


```{r}
# import pandas as pd
# import numpy as np
# 
# # Define the path to the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/subdataset_for_15_perc.parquet"
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# df = pd.read_parquet(input_file)
# 
# # Convert all values to numeric, coerce errors to NaN
# print("Converting values to numeric...")
# df = df.apply(pd.to_numeric, errors='coerce')
# 
# # Flatten the DataFrame to a single array, ignoring NaN values
# all_values = df.values.flatten()
# all_values = all_values[~np.isnan(all_values)]
# 
# # Calculate the 15th percentile value
# percentile_15th = np.percentile(all_values, 15)
# 
# print(f"The 15th percentile value is: {percentile_15th}")

```


The 15th percentile value is: 0.0347440141330495. Now we want to remove those genes where more than 85% (447) of the patients' beta_value is less than 0.0347440141330495, a new column will be created to see how many patients' beta_value is less than 0.0347440141330495.


```{python}
# import pandas as pd
# import numpy as np
# 
# # Define the input and output file paths
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet"
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methy_calculated_patient.parquet"
# 
# # Define the 15th percentile value
# percentile_15th_value = 0.0347440141330495
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# df = pd.read_parquet(input_file)
# 
# # Initialize the new column with NaN values
# df.insert(5, "Patients_Beta_Value_Count", np.nan)
# 
# # Find the starting column index where the header is "tcga-4p-aa8j"
# start_col_index = df.columns.get_loc("tcga-4p-aa8j")
# 
# # Calculate the count of beta_values <= percentile_15th_value for each row (starting from the second row)
# print("Calculating the number of patients' beta_value <= 0.0347440141330495...")
# for index, row in df.iterrows():
#     if index > 0:  # Skip the first row
#         beta_values = row[start_col_index:]
#         # Convert beta_values to numeric for comparison, keep as strings in the DataFrame
#         count = (pd.to_numeric(beta_values, errors='coerce') <= percentile_15th_value).sum()
#         df.at[index, "Patients_Beta_Value_Count"] = count
# 
# # Save the updated DataFrame to a new Parquet file
# print("Saving the updated DataFrame to a new Parquet file...")
# df.to_parquet(output_file, index=False)
```



Check how many rows could be removed.


```{r}
methyl_calculated_patient <- open_dataset("methy_calculated_patient.parquet")
methyl_calculated_patient <- as.data.frame(methyl_calculated_patient)

summary(methyl_calculated_patient$Patients_Beta_Value_Count >= 447)
```



There will be 36689 genes removed, 359376 genes left, 526 patients in total.


```{r remove genes and update calculation}
# import pandas as pd
# import numpy as np
# 
# # Define the input and output file paths
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methy_calculated_patient.parquet"
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_removed_genes.parquet"
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# df = pd.read_parquet(input_file)
# 
# # Replace string 'nan' with actual NaN values
# df.replace('nan', np.nan, inplace=True)
# 
# # Preserve the first row (Summation)
# summation_row = df.iloc[0].copy()
# 
# # Remove rows where Patients_Beta_Value_Count >= 447, excluding the first row
# print("Removing rows where Patients_Beta_Value_Count >= 447...")
# df = df.iloc[1:]
# df = df[df["Patients_Beta_Value_Count"] < 447]
# 
# # Reinsert the Summation row at the top
# df = pd.concat([summation_row.to_frame().T, df]).reset_index(drop=True)
# 
# # Recalculate the first row "Summation"
# print("Recalculating the Summation row...")
# 
# # Specify the columns for summation (4th, 5th, and 7th onward are patients)
# columns_to_summate = ['Low', 'High'] + list(df.columns[6:])
# 
# # Convert the specified columns to numeric for summation, ignoring NaN values
# summation_values = df.loc[1:, columns_to_summate].apply(pd.to_numeric, errors='coerce').sum(axis=0, skipna=True)
# 
# # Convert the summation values back to strings with full decimals
# summation_values_str = summation_values.apply(lambda x: f"{x:.16f}")
# 
# # Update the Summation row in the DataFrame
# df.loc[0, columns_to_summate] = summation_values_str
# 
# # Find the Low and High values for the Summation row
# low_value = summation_values.min()
# high_value = summation_values.max()
# 
# # Update the Low and High columns in the Summation row
# df.loc[0, 'Low'] = f"{low_value:.16f}"
# df.loc[0, 'High'] = f"{high_value:.16f}"
# 
# # Save the updated DataFrame to a new Parquet file
# print("Saving the updated DataFrame to a new Parquet file...")
# df.to_parquet(output_file, index=False)
# 
# print("All operations completed and new Parquet file saved.")

# /Users/yufeiwang/Desktop/HNSCC/methyl_remove_genes.py:24: FutureWarning: The behavior of DataFrame concatenation with empty or all-NA entries is deprecated. In a future version, this will no longer exclude empty or all-NA columns when determining the result dtypes. To retain the old behavior, exclude the relevant entries before the concat operation.
#   df = pd.concat([summation_row.to_frame().T, df]).reset_index(drop=True)

```




```{r add new summation columm to the clinical data}
methyl_removed_genes <- open_dataset("methyl_removed_genes.parquet")
methyl_removed_genes <- as.data.frame(methyl_removed_genes)

# Ready for binding new summation values
clinical_focused_combined_new <- clinical_focused_combined %>% 
  select(-Summation)

clinical_focused_combined_new <- clinical_focused_combined_new[order(rownames(clinical_focused_combined_new)), ]

# Read the CSV file
methyl_sum_row_new <- fread("extracted_first_row_new.csv", stringsAsFactors = FALSE)

# Transpose the data
methyl_sum_row_new <- t(methyl_sum_row_new)

# Set column name
colnames(methyl_sum_row_new) <- "Summation"

# Sort rows by their row names
sorted_methyl_sum_new <- methyl_sum_row_new[order(rownames(methyl_sum_row_new)), ]

# Convert to data frame
sorted_methyl_sum_new <- as.data.frame(sorted_methyl_sum_new, stringsAsFactors = FALSE)

# Set column name
colnames(sorted_methyl_sum_new) <- "Summation"

# Ensure the Summation column retains all decimal places
sorted_methyl_sum_new$Summation <- format(as.numeric(sorted_methyl_sum_new$Summation), digits = 16, scientific = FALSE)

same_names_new <- all(rownames(clinical_focused_combined_new) == rownames(sorted_methyl_sum_new))

if (same_names_new) {
  clinical_focused_combined_new <- cbind(sorted_methyl_sum_new, clinical_focused_combined_new)
} else {
  print("The row names of the datasets are not the same or not in the same order.")
}

clinical_focused_combined_new$Summation <- as.numeric(clinical_focused_combined_new$Summation)

# Create the box plot
ggplot(clinical_focused_combined_new, aes(x = hpv_status, y = Summation)) +
  geom_boxplot() +
  labs(title = "Box Plot of Beta_value Summation by HPV Status",
       x = "HPV Status",
       y = "Beta-value Summation") +
  theme_minimal()

# # Create the box plot with jittered dots
# ggplot(clinical_focused_combined_new, aes(x = hpv_status, y = Summation)) +
#   geom_boxplot(outlier.shape = NA) +  # Hide the default outliers to avoid overlapping with jittered points
#   geom_jitter(width = 0.2, alpha = 0.5, color = "blue") +
#   labs(title = "Box Plot of Beta-value Summation by HPV Status",
#        x = "HPV Status",
#        y = "Beta-value Summation") +
#   theme_minimal()

# Save the plot
ggsave("boxplot_summation_by_hpv_status_after_removal.png")

# Save the cleaned dataframe to a new csv file
write_csv(clinical_focused_combined_new, "clinical_focused_combined_new.csv")
```



```{r t-test after removing genes}
clinical_ttest <- fread("clinical_focused_combined_new.csv")
clinical_ttest <- as.data.frame(clinical_ttest)

# Perform t-test between HPV positive and HPV negative groups
t_test_result_new <- t.test(Summation ~ hpv_status, data = clinical_ttest)

print(t_test_result_new)
```



```{r wilcoxon test after removal}
wilcox_test <- wilcox.test(Summation ~ hpv_status, data = clinical_ttest, exact = FALSE)

print(wilcox_test)
```


```{r Table of anatomic neoplasm subdivision and HPV status after cleaning}
subdivision_new <- as.factor(clinical_ttest$anatomic_neoplasm_subdivision)

is_hpv_new <- as.factor(clinical_ttest$hpv_status)

table1_new <- table(subdivision_new, is_hpv_new)

table1_new_margin <- addmargins(table1_new)

kable(table1_new_margin, format = "html",
      caption = "Cross-Tabulation of Anatomic Neoplasm Subdivision and HPV Status") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                    full_width = FALSE)
```



The subdivision "tonsil" has 7 negative and 38 positive for hpv_status, now we perform a t-test for this specific subdivision.



```{r t-test for tonsil subdivision}
# Filter the data for anatomic_neoplasm_subdivision == "tonsil"
clinical_ttest_filtered <- clinical_ttest %>%
  filter(anatomic_neoplasm_subdivision == "tonsil")

# Check for any conversion warnings
if (any(is.na(clinical_ttest_filtered$Summation))) {
  warning("There are NA values in the Summation column.")
}

# Perform t-test
t_test_result_tonsil <- t.test(Summation ~ hpv_status, data = clinical_ttest_filtered)

# Print the test result
print(t_test_result_tonsil)
```


```{r plot for tonsil subdivision}
# Create the box plot with jittered dots
ggplot(clinical_ttest_filtered, aes(x = hpv_status, y = Summation)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.5, color = "blue") +  
  labs(title = "Box Plot of Beta-value Summation by HPV Status",
       x = "HPV Status",
       y = "Beta-value Summation") +
  theme_minimal()
```



We will calculate three different epigenetic clocks: Horvath, Hannum, and PhenoAge.



```{r Install packages}
# BiocManager::install(c("ENmix", "minfi", "wateRmelon", "FlowSorted.Blood.EPIC", "FlowSorted.Blood.450k", "impute"))
```


Cleaned the methylation dataset to meet the requirements of the format.



```{python Remove unnecessary rows and columns for calculating clocks}
# import pandas as pd
# 
# # Define the input and output file paths
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_removed_genes.parquet"
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_clock.parquet"
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# df = pd.read_parquet(input_file, engine='pyarrow')
# 
# # Remove the first row
# df = df.iloc[1:]
# 
# # Remove columns 2 to 6 
# df = df.drop(df.columns[1:6], axis=1)
# 
# # Save the updated DataFrame to a new Parquet file
# print("Saving the updated DataFrame to a new Parquet file...")
# df.to_parquet(output_file, index=False, engine='pyarrow')
# 
# print("All operations completed and new Parquet file saved.")
```



```{python Impute NA values}
# import pandas as pd
# import numpy as np
# from sklearn.impute import SimpleImputer
# import pyarrow as pa
# import pyarrow.parquet as pq
# 
# # Define the input and output file paths
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_clock.parquet"
# output_matrix_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_imputed_matrix.parquet"
# output_row_names_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_row_names.parquet"
# 
# # Read the Parquet file into a DataFrame
# print("Reading Parquet file...")
# df = pd.read_parquet(input_file)
# 
# # Extract CpG identifiers
# cpg_ids = df['Hybridization REF'].values
# 
# # Convert relevant columns to numeric, coercing errors to NaN
# beta_values = df.iloc[:, 1:].apply(pd.to_numeric, errors='coerce')
# 
# # Impute missing values using mean imputation
# print("Imputing missing values using mean imputation...")
# imputer = SimpleImputer(strategy='mean')
# beta_values_imputed = imputer.fit_transform(beta_values)
# 
# # Create DataFrame from imputed matrix
# imputed_df = pd.DataFrame(beta_values_imputed, columns=df.columns[1:])
# 
# # Reattach the first column (CpG identifiers)
# imputed_df.insert(0, 'Hybridization REF', cpg_ids)
# 
# # Save the imputed DataFrame to a new Parquet file
# print("Saving the imputed DataFrame to a new Parquet file...")
# imputed_df.to_parquet(output_matrix_file, index=False)
# 
# # Save the row names separately to Parquet
# print("Saving the row names to a new Parquet file...")
# row_names_df = pd.DataFrame({'Hybridization REF': cpg_ids})
# row_names_df.to_parquet(output_row_names_file, index=False)
# 
# print("All operations completed and new Parquet files saved.")
```



```{r}
# BiocManager::install("doParallel")
# BiocManager::install("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
```




At this point, I realize that for specific clocks, we need to use those specific CpGs listed in the data object "age_coefficient", we need to roll back to the place where CpGs were not deleted (10th percentile), and filter out those CpGs for each specific clock. Refer to chunk 27, the file is called "methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet"




```{r Extract coefficient lists for different clocks}
data(age_coefficients)

horvath_cpgs <- setdiff(names(ageCoefs$Horvath), "(Intercept)")
hannum_cpgs <- names(ageCoefs$Hannum)
pheno_cpgs <- setdiff(names(ageCoefs$PhenoAge), "(Intercept)")

write.csv(horvath_cpgs, file = "horvath_cpgs.csv", row.names = FALSE)
write.csv(hannum_cpgs, file = "hannum_cpgs.csv", row.names = FALSE)
write.csv(pheno_cpgs, file = "pheno_cpgs.csv", row.names = FALSE)
```



```{python Filter out rows}
# import pandas as pd
# import pyarrow.parquet as pq
# from pyarrow import Table
# from tqdm import tqdm
# 
# # Load the methylation dataset
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet"
# df = pd.read_parquet(input_file)
# 
# # Step 1: Remove the first row
# df = df.iloc[1:].reset_index(drop=True)
# 
# # Step 2: Remove columns 2-5
# df.drop(df.columns[1:5], axis=1, inplace=True)
# 
# # Function to check and print CSV columns
# def check_csv_columns(file_path):
#     df_csv = pd.read_csv(file_path)
#     print(f"Columns in {file_path}: {df_csv.columns.tolist()}")
#     return df_csv
# 
# # Check the columns in the CSV files
# horvath_cpgs_df = check_csv_columns("/Users/yufeiwang/Desktop/HNSCC/horvath_cpgs.csv")
# hannum_cpgs_df = check_csv_columns("/Users/yufeiwang/Desktop/HNSCC/hannum_cpgs.csv")
# pheno_cpgs_df = check_csv_columns("/Users/yufeiwang/Desktop/HNSCC/pheno_cpgs.csv")
# 
# # Extract the CpG lists
# horvath_cpgs = horvath_cpgs_df.iloc[:, 0].tolist()
# hannum_cpgs = hannum_cpgs_df.iloc[:, 0].tolist()
# pheno_cpgs = pheno_cpgs_df.iloc[:, 0].tolist()
# 
# # Function to filter and save dataset
# def filter_and_save(df, cpg_list, cpg_name, output_file):
#     missing_cpgs = [cpg for cpg in cpg_list if cpg not in df['Hybridization REF'].values]
#     filtered_df = df[df['Hybridization REF'].isin(cpg_list)]
#     
#     if missing_cpgs:
#         print(f"Missing CpGs in {cpg_name}: {missing_cpgs}")
#     
#     filtered_data_arrow = Table.from_pandas(filtered_df)
#     pq.write_table(filtered_data_arrow, output_file)
#     print(f"Filtered dataset for {cpg_name} saved to {output_file}")
# 
# # Filter and save the datasets
# filter_and_save(df, horvath_cpgs, "Horvath", "/Users/yufeiwang/Desktop/HNSCC/filtered_methylation_data_horvath.parquet")
# filter_and_save(df, hannum_cpgs, "Hannum", "/Users/yufeiwang/Desktop/HNSCC/filtered_methylation_data_hannum.parquet")
# filter_and_save(df, pheno_cpgs, "PhenoAge", "/Users/yufeiwang/Desktop/HNSCC/filtered_methylation_data_phenoage.parquet")

```




```{r Calculate clocks but found lots of missing CpGs}
horvath_data <- read_parquet("filtered_methylation_data_horvath.parquet")
hannum_data <- read_parquet("filtered_methylation_data_hannum.parquet")
phenoage_data <- read_parquet("filtered_methylation_data_phenoage.parquet")

# Delete the last column
horvath_data <- horvath_data[, -ncol(horvath_data)]
hannum_data <- hannum_data[, -ncol(hannum_data)]
phenoage_data <- phenoage_data[, -ncol(phenoage_data)]


# Function to convert data frame to numeric matrix
convert_to_numeric_matrix <- function(data) {
  if (ncol(data) < 2) {
    stop("Data must have at least two columns")
  }
  
  cpg_ids <- data[[1]]  # Extract the first column as CpG identifiers
  data <- data[, -1]  # Exclude the first column
  
  # Check for unique CpG identifiers
  if (length(unique(cpg_ids)) != length(cpg_ids)) {
    stop("CpG identifiers are not unique")
  }
  
  # Ensure all values are numeric while preserving row names
  data_matrix <- apply(data, 2, function(x) as.numeric(as.character(x)))
  rownames(data_matrix) <- cpg_ids
  
  if (length(rownames(data_matrix)) != nrow(data_matrix)) {
    print(paste("Number of row names:", length(rownames(data_matrix))))
    print(paste("Number of rows:", nrow(data_matrix)))
    stop("Mismatch between number of row names and number of rows")
  }
  return(data_matrix)
}

# Convert data to numeric matrices
horvath_betas <- convert_to_numeric_matrix(horvath_data)
hannum_betas <- convert_to_numeric_matrix(hannum_data)
phenoage_betas <- convert_to_numeric_matrix(phenoage_data)

# Check if the matrices contain any non-numeric values
check_numeric <- function(matrix_data, name) {
  if (!is.matrix(matrix_data) || mode(matrix_data) != "numeric") {
    stop(paste("Matrix for", name, "is not numeric"))
  }
}

check_numeric(horvath_betas, "Horvath")
check_numeric(hannum_betas, "Hannum")
check_numeric(phenoage_betas, "PhenoAge")

# Calculate the epigenetic clocks
horvath_age <- agep(horvath_betas, method = "horvath")
hannum_age <- agep(hannum_betas, method = "hannum")
phenoage_age <- agep(phenoage_betas, method = "phenoage")

# Check if row names are the same
same_rownames <- all(rownames(horvath_age) == rownames(hannum_age)) && all(rownames(horvath_age) == rownames(phenoage_age))

if (same_rownames) {
  # Combine results into a data frame with rownames
  clock_ages <- data.frame(Horvath = horvath_age[,1], Hannum = hannum_age[,1], PhenoAge = phenoage_age[,1])
  rownames(clock_ages) <- rownames(horvath_age)
  print(clock_ages)
} else {
  print("Row names do not match between the age calculations.")
}
```


Now we roll back to the original data again to impute the NA value with the median value of the column, then calculate the clocks again. "methyl_calculated_removed_indeter_hpv.parquet" is used.


```{r Take a look at the original data}
ori_data <- read_parquet("methyl_calculated_removed_indeter_hpv.parquet")
ori_data <- as.data.frame(ori_data)
```




```{python Impute NA and filter out rows}
# import pandas as pd
# import pyarrow.parquet as pq
# from pyarrow import Table
# 
# # Load the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_calculated_removed_indeter_hpv.parquet"
# df = pd.read_parquet(input_file)
# 
# # Remove the first row and columns 2 to 5
# df = df.iloc[1:]  # Remove the first row
# df = df.drop(df.columns[1:5], axis=1)  # Remove columns 2 to 5
# 
# # Convert beta-values to numeric, keeping all decimals, and impute "nan" with the minimum value of the respective column
# for col in df.columns[1:]:  # Skip the first column as it contains CpG names
#     df[col] = pd.to_numeric(df[col], errors='coerce')
#     min_value = df[col].min(skipna=True)
#     df[col] = df[col].fillna(min_value)
# 
# # Add a new row called "Summation" as the first row
# summation_row = df.iloc[:, 1:].sum()
# summation_df = pd.DataFrame([summation_row], columns=df.columns[1:])
# summation_df.insert(0, 'Hybridization REF', 'Summation')
# df = pd.concat([summation_df, df]).reset_index(drop=True)
# 
# # Ensure no scientific notation
# pd.set_option('display.float_format', lambda x: f'{x:.16f}')
# 
# # Save the entire methylation data into a Parquet file
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_imputed_sum.parquet"
# df_arrow = Table.from_pandas(df)
# pq.write_table(df_arrow, output_file)
# print(f"Entire methylation data saved to {output_file}")
# 
# # Extract the summation row into a CSV file without the "Hybridization REF" column
# summation_row_with_header = summation_row.to_frame().T
# summation_row_with_header.index = ['Summation']
# 
# # Save the CSV file with full decimal precision
# summation_row_with_header.to_csv("/Users/yufeiwang/Desktop/HNSCC/new_summation_row.csv", index=True, header=True, float_format='%.16f')
# print("Summation row saved to new_summation_row.csv")
# 
# # Function to filter and save dataset
# def filter_and_save(df, cpg_list, cpg_name, output_file):
#     missing_cpgs = [cpg for cpg in cpg_list if cpg not in df['Hybridization REF'].values]
#     filtered_df = df[df['Hybridization REF'].isin(cpg_list)]
#     
#     if missing_cpgs:
#         print(f"Missing CpGs in {cpg_name}: {missing_cpgs}")
#     
#     filtered_data_arrow = Table.from_pandas(filtered_df)
#     pq.write_table(filtered_data_arrow, output_file)
#     print(f"Filtered dataset for {cpg_name} saved to {output_file}")
# 
# # Load the CpG lists from CSV files
# horvath_cpgs_df = pd.read_csv("/Users/yufeiwang/Desktop/HNSCC/horvath_cpgs.csv")
# hannum_cpgs_df = pd.read_csv("/Users/yufeiwang/Desktop/HNSCC/hannum_cpgs.csv")
# pheno_cpgs_df = pd.read_csv("/Users/yufeiwang/Desktop/HNSCC/pheno_cpgs.csv")
# 
# # Extract the CpG lists
# horvath_cpgs = horvath_cpgs_df.iloc[:, 0].tolist()
# hannum_cpgs = hannum_cpgs_df.iloc[:, 0].tolist()
# pheno_cpgs = pheno_cpgs_df.iloc[:, 0].tolist()
# 
# # Filter and save the datasets for each clock
# filter_and_save(df, horvath_cpgs, "Horvath", "/Users/yufeiwang/Desktop/HNSCC/filtered_methyl_horvath_new.parquet")
# filter_and_save(df, hannum_cpgs, "Hannum", "/Users/yufeiwang/Desktop/HNSCC/filtered_methyl_hannum_new.parquet")
# filter_and_save(df, pheno_cpgs, "PhenoAge", "/Users/yufeiwang/Desktop/HNSCC/filtered_methyl_phenoage_new.parquet")


```




```{r calculate clocks}
horvath_data_new <- read_parquet("filtered_methyl_horvath_new.parquet")
hannum_data_new <- read_parquet("filtered_methyl_hannum_new.parquet")
phenoage_data_new <- read_parquet("filtered_methyl_phenoage_new.parquet")

# Delete the last column
horvath_data_new <- horvath_data_new[, -ncol(horvath_data_new)]
hannum_data_new <- hannum_data_new[, -ncol(hannum_data_new)]
phenoage_data_new <- phenoage_data_new[, -ncol(phenoage_data_new)]


# Function to convert data frame to numeric matrix
convert_to_numeric_matrix <- function(data) {
  if (ncol(data) < 2) {
    stop("Data must have at least two columns")
  }
  
  cpg_ids <- data[[1]]  # Extract the first column as CpG identifiers
  data <- data[, -1]  # Exclude the first column
  
  # Check for unique CpG identifiers
  if (length(unique(cpg_ids)) != length(cpg_ids)) {
    stop("CpG identifiers are not unique")
  }
  
  # Ensure all values are numeric while preserving row names
  data_matrix <- apply(data, 2, function(x) as.numeric(as.character(x)))
  rownames(data_matrix) <- cpg_ids
  
  if (length(rownames(data_matrix)) != nrow(data_matrix)) {
    print(paste("Number of row names:", length(rownames(data_matrix))))
    print(paste("Number of rows:", nrow(data_matrix)))
    stop("Mismatch between number of row names and number of rows")
  }
  
  
  return(data_matrix)
}

# Convert data to numeric matrices
horvath_betas_new <- convert_to_numeric_matrix(horvath_data_new)
hannum_betas_new <- convert_to_numeric_matrix(hannum_data_new)
phenoage_betas_new <- convert_to_numeric_matrix(phenoage_data_new)

# Check if the matrices contain any non-numeric values
check_numeric <- function(matrix_data, name) {
  if (!is.matrix(matrix_data) || mode(matrix_data) != "numeric") {
    stop(paste("Matrix for", name, "is not numeric"))
  }
}

check_numeric(horvath_betas_new, "Horvath")
check_numeric(hannum_betas_new, "Hannum")
check_numeric(phenoage_betas_new, "PhenoAge")

# Calculate Horvath Clock
# horvath_age_new <- agep(horvath_betas_new, method = "horvath", n_missing = TRUE)
# print(horvath_age_new)

# Calculate Hannum Clock
# hannum_age_new <- agep(hannum_betas_new, method = "hannum", n_missing = TRUE)
# print(hannum_age_new)

# Calculate PhenoAge
# phenoage_new <- agep(phenoage_betas_new, method = "phenoage", n_missing = TRUE)
# print(phenoage_new)

# Calculate the epigenetic clocks
horvath_age <- agep(horvath_betas_new, method = "horvath")
hannum_age <- agep(hannum_betas_new, method = "hannum")
phenoage_age <- agep(phenoage_betas_new, method = "phenoage")

# Check if row names are the same
same_rownames <- all(rownames(horvath_age) == rownames(hannum_age)) && all(rownames(horvath_age) == rownames(phenoage_age))

if (same_rownames) {
  # Combine results into a data frame with rownames
  clock_ages <- data.frame(Horvath = horvath_age[,1], Hannum = hannum_age[,1], PhenoAge = phenoage_age[,1])
  rownames(clock_ages) <- rownames(horvath_age)
  print(clock_ages)
} else {
  print("Row names do not match between the age calculations.")
}

```



After calculating three clocks, I will merge them into clinical data and then do the age difference test. For methylation data, I will calculate the summation value of all CpGs island beta-value based on the imputed data.


```{r Glimpse of the new summation row and the methylation data with imputation}
new_sum_row <- read.csv("new_summation_row.csv")

methyl_imputed_new <- read_parquet("methyl_imputed_sum.parquet")
methyl_imputed_new <- as.data.frame(methyl_imputed_new)
```



```{r Load the clinical data again}
# Transpose the clinical data
clinical_transposed <- t(clinical_all_CDEs)

clinical_transposed <- as.data.frame(clinical_transposed)

colnames(clinical_transposed) <- clinical_transposed[1,]

clinical_transposed <- clinical_transposed[-1,]

# Select necessary variables
clinical_focused_for_age <- clinical_transposed[, c(9, 12, 21:23, 46, 48, 54, 114)]

# Filter data
clinical_focused_filtered_for_age <- clinical_focused_for_age %>%
  filter(hpv_status != "indeterminate", tumor_tissue_site == "head and neck")
```




```{r Merge clocks and the summation row into clinical data}
# Process the summation row
rownames(new_sum_row) <- new_sum_row[,1]
new_sum_row <- new_sum_row[ , -1]

# Transpose the data
new_sum_row_transpose <- t(new_sum_row)

# Convert to data.frame
new_sum_row_transpose <- as.data.frame(new_sum_row_transpose)

# Sort the row names
sorted_new_sum_row_transpose <- new_sum_row_transpose[order(rownames(new_sum_row_transpose)), , drop = FALSE]
rownames(sorted_new_sum_row_transpose) <- gsub("\\.", "-", rownames(sorted_new_sum_row_transpose))
row_name_sum <- rownames(sorted_new_sum_row_transpose)

# Process clocks data (sort rownames)
sorted_clock_ages <- clock_ages[order(rownames(clock_ages)), , drop = FALSE]
row_name_clock <- rownames(sorted_clock_ages)


# Process clinical data (sort rownames)
sorted_clinical_focused_filtered_for_age <- clinical_focused_filtered_for_age[order(rownames(clinical_focused_filtered_for_age)), , drop = FALSE]
row_name_clinical <- rownames(sorted_clinical_focused_filtered_for_age)



# Merge them into clinical data
all_colnames_match <- all(row_name_sum == row_name_clock)

if (all_colnames_match) {
  # Merge the data frames
  merged_df <- cbind(sorted_new_sum_row_transpose, sorted_clock_ages)
  
  # Further merge with the clinical data
  all_colnames_match_clinical <- all(rownames(merged_df) == row_name_clinical)
  
  if (all_colnames_match_clinical) {
    sorted_clinical_focused_filtered_for_age <- cbind(merged_df, clinical_focused_filtered_for_age)
    
    # Save the merged data frame
    write.csv(sorted_clinical_focused_filtered_for_age, "/Users/yufeiwang/Desktop/HNSCC/sorted_clinical_focused_filtered_for_age.csv", row.names = TRUE)
    print("Merged data frame saved successfully.")
  } else {
    print("The row names of the merged data frame and clinical data frame do not match.")
  }
} else {
  print("The column names of the summation row transposed and clock ages data frames do not match.")
}

```



The merged clinical data is called "sorted_clinical_focused_filtered_for_age".



```{r Filter out age is missing and calculate age difference}
# Filter out age is missing
clinical_filtered_age <- sorted_clinical_focused_filtered_for_age %>% 
  filter(!is.na(age_at_initial_pathologic_diagnosis))

# Ensure age_at_initial_pathologic_diagnosis is numeric
clinical_filtered_age$age_at_initial_pathologic_diagnosis <- as.numeric(clinical_filtered_age$age_at_initial_pathologic_diagnosis)

# Calculate the differences and add them as new columns
clinical_filtered_age$Horvath_diff <- abs(clinical_filtered_age$Horvath - clinical_filtered_age$age_at_initial_pathologic_diagnosis)
clinical_filtered_age$Hannum_diff <- abs(clinical_filtered_age$Hannum - clinical_filtered_age$age_at_initial_pathologic_diagnosis)
clinical_filtered_age$PhenoAge_diff <- abs(clinical_filtered_age$PhenoAge - clinical_filtered_age$age_at_initial_pathologic_diagnosis)

# Reorder columns to place the new columns in the 5th, 6th, and 7th positions
column_order <- c(names(clinical_filtered_age)[1:4], 'Horvath_diff', 'Hannum_diff', 'PhenoAge_diff', names(clinical_filtered_age)[5:(ncol(clinical_filtered_age)-3)])
clinical_filtered_age <- clinical_filtered_age[, column_order]


```



The latest clinical data is called "clinical_filtered_age", which has the most complete information.



```{r Summation test}
# Create the boxplot
ggplot(clinical_filtered_age, aes(x = hpv_status, y = Summation)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Add points for individual data
  labs(title = "Box Plot of Summation by HPV Status",
       x = "HPV Status",
       y = "Summation") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(clinical_filtered_age$Summation)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Summation ~ hpv_status, data = clinical_filtered_age)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Summation ~ hpv_status, data = clinical_filtered_age)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```




```{r Test for Horvath_diff}
# Create the boxplot
ggplot(clinical_filtered_age, aes(x = hpv_status, y = Horvath_diff)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Add points for individual data
  labs(title = "Box Plot of Horvath_diff by HPV Status",
       x = "HPV Status",
       y = "Horvath_diff") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(clinical_filtered_age$Horvath_diff)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Horvath_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Horvath_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```


```{r Horvath clock}
# Create the boxplot
ggplot(clinical_filtered_age, aes(x = hpv_status, y = Horvath)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Add points for individual data
  labs(title = "Box Plot of Horvath by HPV Status",
       x = "HPV Status",
       y = "Horvath_diff") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(clinical_filtered_age$Horvath_diff)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Horvath_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Horvath_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```



```{r Test for Hannum_diff}
# Create the boxplot
ggplot(clinical_filtered_age, aes(x = hpv_status, y = Hannum_diff)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Add points for individual data
  labs(title = "Box Plot of Hannum_diff by HPV Status",
       x = "HPV Status",
       y = "Hannum_diff") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(clinical_filtered_age$Hannum_diff)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Hannum_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Hannum_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```




```{r Test for PhenoAge_diff}
# Create the boxplot
ggplot(clinical_filtered_age, aes(x = hpv_status, y = PhenoAge_diff)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Add points for individual data
  labs(title = "Box Plot of PhenoAge_diff by HPV Status",
       x = "HPV Status",
       y = "PhenoAge_diff") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(clinical_filtered_age$PhenoAge_diff)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(PhenoAge_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(PhenoAge_diff ~ hpv_status, data = clinical_filtered_age)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```



For the data without imputation and with imputation, the test results remain the same.


```{r Test for true age}
# Create the boxplot
ggplot(clinical_filtered_age, aes(x = hpv_status, y = age_at_initial_pathologic_diagnosis)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Add points for individual data
  labs(title = "Box Plot of age by HPV Status",
       x = "HPV Status",
       y = "Age") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(clinical_filtered_age$age_at_initial_pathologic_diagnosis)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(age_at_initial_pathologic_diagnosis ~ hpv_status, data = clinical_filtered_age)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(age_at_initial_pathologic_diagnosis ~ hpv_status, data = clinical_filtered_age)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```


```{r Pair t-test for three clocks and true age}
# HPV+ group
hpv_positive <- clinical_filtered_age %>% filter(hpv_status == "positive")

t_test_hannum_hpv_pos <- t.test(hpv_positive$Hannum, hpv_positive$age_at_initial_pathologic_diagnosis, paired = TRUE)
t_test_phenoage_hpv_pos <- t.test(hpv_positive$PhenoAge, hpv_positive$age_at_initial_pathologic_diagnosis, paired = TRUE)
t_test_horvath_hpv_pos <- t.test(hpv_positive$Horvath, hpv_positive$age_at_initial_pathologic_diagnosis, paired = TRUE)

# HPV- group
hpv_negative <- clinical_filtered_age %>% filter(hpv_status == "negative")

t_test_hannum_hpv_neg <- t.test(hpv_negative$Hannum, hpv_negative$age_at_initial_pathologic_diagnosis, paired = TRUE)
t_test_phenoage_hpv_neg <- t.test(hpv_negative$PhenoAge, hpv_negative$age_at_initial_pathologic_diagnosis, paired = TRUE)
t_test_horvath_hpv_neg <- t.test(hpv_negative$Horvath, hpv_negative$age_at_initial_pathologic_diagnosis, paired = TRUE)

# Print results
print(t_test_hannum_hpv_pos)
print(t_test_phenoage_hpv_pos)
print(t_test_horvath_hpv_pos)

print(t_test_hannum_hpv_neg)
print(t_test_phenoage_hpv_neg)
print(t_test_horvath_hpv_neg)

```





```{r Wilcoxon signed-rank test for three clocks and true age}
# HPV+ group
wilcox_test_hannum_hpv_pos <- wilcox.test(hpv_positive$Hannum, hpv_positive$age_at_initial_pathologic_diagnosis, paired = TRUE)
wilcox_test_phenoage_hpv_pos <- wilcox.test(hpv_positive$PhenoAge, hpv_positive$age_at_initial_pathologic_diagnosis, paired = TRUE)
wilcox_test_horvath_hpv_pos <- wilcox.test(hpv_positive$Horvath, hpv_positive$age_at_initial_pathologic_diagnosis, paired = TRUE)

# HPV- group
wilcox_test_hannum_hpv_neg <- wilcox.test(hpv_negative$Hannum, hpv_negative$age_at_initial_pathologic_diagnosis, paired = TRUE)
wilcox_test_phenoage_hpv_neg <- wilcox.test(hpv_negative$PhenoAge, hpv_negative$age_at_initial_pathologic_diagnosis, paired = TRUE)
wilcox_test_horvath_hpv_neg <- wilcox.test(hpv_negative$Horvath, hpv_negative$age_at_initial_pathologic_diagnosis, paired = TRUE)

# Print results
print(wilcox_test_hannum_hpv_pos)
print(wilcox_test_phenoage_hpv_pos)
print(wilcox_test_horvath_hpv_pos)

print(wilcox_test_hannum_hpv_neg)
print(wilcox_test_phenoage_hpv_neg)
print(wilcox_test_horvath_hpv_neg)

```





Now we find that epigenetic clocks assumptions don't work, we will use both filtered data to focus on the summation for women and men in different HPV groups.


```{r Load the clinical data again}
# Transpose the clinical data
clinical_transposed <- t(clinical_all_CDEs)

clinical_transposed <- as.data.frame(clinical_transposed)

colnames(clinical_transposed) <- clinical_transposed[1,]

clinical_transposed <- clinical_transposed[-1,]

# Select necessary variables
clinical_focused_for_age <- clinical_transposed[, c(9, 12, 21:23, 46, 48, 54, 114)]

# Filter data
clinical_focused_filtered_for_age <- clinical_focused_for_age %>%
  filter(hpv_status != "indeterminate", tumor_tissue_site == "head and neck")
```




```{r Combine two columns of summation with the clinical data}
# only removed nan rows (396066 left)
methyl_sum_row <- fread("first_row_extracted.csv")
methyl_sum_row <- t(methyl_sum_row)
colnames(methyl_sum_row) <- "Summation"
sorted_methyl_sum <- methyl_sum_row[order(rownames(methyl_sum_row)), ]
sorted_methyl_sum <- as.data.frame(sorted_methyl_sum)
colnames(sorted_methyl_sum) <- "Summation"


# further cleaned (359376 left)
methyl_sum_row_new <- fread("extracted_first_row_new.csv")
methyl_sum_row_new <- t(methyl_sum_row_new)
colnames(methyl_sum_row_new) <- "Summation_cleaned"
sorted_methyl_sum_new <- methyl_sum_row_new[order(rownames(methyl_sum_row_new)), ]
sorted_methyl_sum_new <- as.data.frame(sorted_methyl_sum_new)
colnames(sorted_methyl_sum_new) <- "Summation_cleaned"


# Process clinical data (sort rownames)
sorted_clinical_focused_filtered_for_age <- clinical_focused_filtered_for_age[order(rownames(clinical_focused_filtered_for_age)), , drop = FALSE]
row_name_clinical <- rownames(sorted_clinical_focused_filtered_for_age)


# Merge them

# Remove rows in sorted_methyl_sum that are not in row_name_clinical
filtered_methyl_sum <- sorted_methyl_sum[row_name_clinical, , drop = FALSE]

# Check if the row names are consistent
all_row_names_match <- all(rownames(filtered_methyl_sum) == row_name_clinical) &&
                       all(rownames(sorted_methyl_sum_new) == row_name_clinical)
if (all_row_names_match) {
  final_clinical_data <- cbind(filtered_methyl_sum, sorted_methyl_sum_new, sorted_clinical_focused_filtered_for_age)
  print("The row names are consistent and the data has been merged successfully.")
} else {
  print("The row names do not match. Please check the data.")
}

```





```{r Summation test (396066) 1A1}
# Create the boxplot (title = "Box Plot of Global Methylation by HPV Status")
figure_1A <- ggplot(final_clinical_data, aes(x = hpv_status, y = Summation)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(x = "HPV Status",
       y = "Global Methylation") +
  theme_minimal()

# Save the plot as a PNG file with 300 DPI
ggsave("1A-boxplot_global_methylation_by_hpv_status.png", plot = figure_1A, width = 8, height = 6, dpi = 300)

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(final_clinical_data$Summation)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Summation ~ hpv_status, data = final_clinical_data)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Summation ~ hpv_status, data = final_clinical_data)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```



```{r Summation_cleaned test (359376)}
# Create the boxplot
ggplot(final_clinical_data, aes(x = hpv_status, y = Summation_cleaned)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(title = "Box Plot of Summation_cleaned by HPV Status",
       x = "HPV Status",
       y = "Summation_cleaned") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test)
shapiro_test <- shapiro.test(final_clinical_data$Summation_cleaned)

if (shapiro_test$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Summation_cleaned ~ hpv_status, data = final_clinical_data)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Summation_cleaned ~ hpv_status, data = final_clinical_data)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```


$$NOTE: this\ section\ below\ is\ done\ when\ preparing\ for\ the\ poster.$$



```{r Divide gender group}
# male group
male_group <- final_clinical_data %>% filter(gender == "male")

# female group
female_group <- final_clinical_data %>% filter(gender == "female")
```




```{r Box plot for male group}
# Create the boxplot (title = "Box Plot of Global Methylation by HPV Status in Male Group")
figure_1B1 <- ggplot(male_group, aes(x = hpv_status, y = Summation)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(x = "HPV Status",
       y = "Global Methylation") +
  theme_minimal()

# Save the plot as a PNG file with 300 DPI
ggsave("1B1-boxplot_global_methylation_by_hpv_status_male_group.png", plot = figure_1B1, width = 8, height = 6, dpi = 300)
```




```{r Box plot for female group}
# Create the boxplot (title = "Box Plot of Global Methylation by HPV Status in Female Group")
figure_1B2 <- ggplot(female_group, aes(x = hpv_status, y = Summation)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(x = "HPV Status",
       y = "Global Methylation") +
  theme_minimal()

# Save the plot as a PNG file with 300 DPI
ggsave("1B2-boxplot_global_methylation_by_hpv_status_female_group.png", plot = figure_1B2, width = 8, height = 6, dpi = 300)
```



$$END\ NOTE$$



```{r Divide HPV groups}
# HPV+ group
hpv_positive <- final_clinical_data %>% filter(hpv_status == "positive")

# HPV- group
hpv_negative <- final_clinical_data %>% filter(hpv_status == "negative")
```




```{r Summation test by gender in HPV positive group}
# Create the boxplot for the HPV+ group
ggplot(hpv_positive, aes(x = gender, y = Summation)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(title = "Box Plot of Summation by Gender in HPV positive group",
       x = "Gender",
       y = "Summation") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test) for each gender group separately
shapiro_test_male <- shapiro.test(hpv_positive %>% filter(gender == "male") %>% pull(Summation))
shapiro_test_female <- shapiro.test(hpv_positive %>% filter(gender == "female") %>% pull(Summation))

# Use Wilcoxon rank-sum test if either group is not normally distributed
if (shapiro_test_male$p.value < 0.05 | shapiro_test_female$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Summation ~ gender, data = hpv_positive)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Summation ~ gender, data = hpv_positive)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```





```{r Summation test by gender in HPV negative group}
# Create the boxplot for the HPV+ group
ggplot(hpv_negative, aes(x = gender, y = Summation)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(title = "Box Plot of Summation by Gender in HPV negative group",
       x = "Gender",
       y = "Summation") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test) for each gender group separately
shapiro_test_male <- shapiro.test(hpv_negative %>% filter(gender == "male") %>% pull(Summation))
shapiro_test_female <- shapiro.test(hpv_negative %>% filter(gender == "female") %>% pull(Summation))

# Use Wilcoxon rank-sum test if either group is not normally distributed
if (shapiro_test_male$p.value < 0.05 | shapiro_test_female$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Summation ~ gender, data = hpv_negative)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Summation ~ gender, data = hpv_negative)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```





```{r Summation_cleaned test by gender in HPV positive group}
# Create the boxplot for the HPV+ group
ggplot(hpv_positive, aes(x = gender, y = Summation_cleaned)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(title = "Box Plot of Summation_cleaned by Gender in HPV positive group",
       x = "Gender",
       y = "Summation_cleaned") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test) for each gender group separately
shapiro_test_male <- shapiro.test(hpv_positive %>% filter(gender == "male") %>% pull(Summation_cleaned))
shapiro_test_female <- shapiro.test(hpv_positive %>% filter(gender == "female") %>% pull(Summation_cleaned))

# Use Wilcoxon rank-sum test if either group is not normally distributed
if (shapiro_test_male$p.value < 0.05 | shapiro_test_female$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Summation_cleaned ~ gender, data = hpv_positive)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Summation_cleaned ~ gender, data = hpv_positive)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```



```{r Summation_cleaned test by gender in HPV negative group}
# Create the boxplot for the HPV+ group
ggplot(hpv_negative, aes(x = gender, y = Summation_cleaned)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +  
  labs(title = "Box Plot of Summation_cleaned by Gender in HPV negative group",
       x = "Gender",
       y = "Summation_cleaned") +
  theme_minimal()

# Check for normality (Shapiro-Wilk test) for each gender group separately
shapiro_test_male <- shapiro.test(hpv_negative %>% filter(gender == "male") %>% pull(Summation_cleaned))
shapiro_test_female <- shapiro.test(hpv_negative %>% filter(gender == "female") %>% pull(Summation_cleaned))

# Use Wilcoxon rank-sum test if either group is not normally distributed
if (shapiro_test_male$p.value < 0.05 | shapiro_test_female$p.value < 0.05) {
  # Data is not normally distributed, use Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(Summation_cleaned ~ gender, data = hpv_negative)
  test_result <- wilcox_test
  test_name <- "Wilcoxon rank-sum test"
} else {
  # Data is normally distributed, use t-test
  t_test <- t.test(Summation_cleaned ~ gender, data = hpv_negative)
  test_result <- t_test
  test_name <- "t-test"
}

# Print the results of the test
cat("Result of the", test_name, ":\n")
print(test_result)
```


Now we focus on genes, we will use the cleaned methylation data.



```{r The original methylation data}
ori_methyl <- read_parquet("methyl_calculated_removed_indeter_hpv.parquet")
ori_methyl <- as.data.frame(ori_methyl)
```



```{r Check number of genes of the original methylation data}
gene_symbol_counts <- ori_methyl %>%
  group_by(Gene_Symbol) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(gene_symbol_counts)
```




```{r The methylation data after cleaning}
# This is the data that have been cleaned for both hpv "indeterminate" and "nan" rows
methyl_merge_calculated_cleaned <- open_dataset("methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet")
methyl_merge_calculated_cleaned <- as.data.frame(methyl_merge_calculated_cleaned)
```



```{r Check number of genes of the clean methylation data}
methyl_merge_calculated_cleaned %>%
  group_by(Gene_Symbol) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```



```{python process filter out genes and save to a parquet file}
# import pandas as pd
# import pyarrow as pa
# import pyarrow.parquet as pq
# 
# # Load the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet"
# df = pd.read_parquet(input_file)
# 
# # Remove the first row (not headers)
# df = df.iloc[1:, :]
# 
# # Remove the 1st, 3rd, 4th, and 5th columns
# df = df.drop(columns=[df.columns[0], df.columns[2], df.columns[3], df.columns[4]])
# 
# # Rename columns for clarity
# df.columns = ['Gene_Symbol'] + df.columns[1:].tolist()
# 
# Remove rows where 'Gene_Symbol' is 'nan' or empty
# df = df[df['Gene_Symbol'].notna() & df['Gene_Symbol'].str.strip().astype(bool) & (df['Gene_Symbol'] != 'nan')]

# # Convert beta-values to numeric, keeping all decimals
# for col in df.columns[1:]:
#     df[col] = pd.to_numeric(df[col], errors='coerce')
# 
# # Sum beta-values for each gene across multiple rows
# df_grouped = df.groupby('Gene_Symbol').sum()
# 
# # Reset the index to have 'Gene_Symbol' as a column again
# df_grouped.reset_index(inplace=True)
# 
# # Save the resulting dataset to a new Parquet file
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_summed_genes.parquet"
# table = pa.Table.from_pandas(df_grouped)
# pq.write_table(table, output_file)
# 
# print(f"Processed data saved to {output_file}")

```



"methyl_calculated_removed_indeter_hpv_and_missing_beta.parquet" is used to clean the methylation data to only contain necessary columns and rows. "nan" or empty value for gene will be removed. Each gene now has only one row, the value is the summation across multiple rows for that gene.



```{r The data that only has unique genes}
methyl_gene <- open_dataset("methyl_summed_genes.parquet")
methyl_gene <- as.data.frame(methyl_gene)
```



```{python Apply threshold to clean genes}
# import pandas as pd
# import pyarrow as pa
# import pyarrow.parquet as pq
# import numpy as np
# 
# # Load the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_summed_genes.parquet"
# df = pd.read_parquet(input_file)
# 
# # Calculate the 10th percentile among all beta-values (excluding the first column)
# beta_values = df.iloc[:, 1:].values.flatten()
# percentile_10th_value = np.percentile(beta_values, 10)
# print(f"10th percentile value: {percentile_10th_value}")
# 
# # Filter out genes where >= 473 patients' beta-values are less than the 10th percentile value
# def filter_genes(row, threshold, min_patients):
#     return (row[1:] < threshold).sum() < min_patients
# 
# filtered_df = df[df.apply(filter_genes, axis=1, args=(percentile_10th_value, 473))]
# 
# # Save the resulting dataset to a new Parquet file
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_cleaned_genes.parquet"
# table = pa.Table.from_pandas(filtered_df)
# pq.write_table(table, output_file)
# 
# print(f"Filtered data saved to {output_file}")


```


The 10th percentile value: 0.5131194491438932. We use 10% patient threshold. 



```{r The data after cleaning genes}
methyl_gene_cleaned <- open_dataset("methyl_cleaned_genes.parquet")
methyl_gene_cleaned <- as.data.frame(methyl_gene_cleaned)

methyl_gene_cleaned <- methyl_gene_cleaned[, -ncol(methyl_gene_cleaned)]
```




```{r Transpose the data to merge into the clinical data}
transposed_methyl_gene_cleaned <- t(methyl_gene_cleaned[, -1])

# Convert it to a data frame
transposed_methyl_gene_cleaned <- as.data.frame(transposed_methyl_gene_cleaned)

# Set the column names to the values in the first row (Gene Symbols)
colnames(transposed_methyl_gene_cleaned) <- methyl_gene_cleaned$Gene_Symbol

# Set the row names to the original column names (Patient IDs)
rownames(transposed_methyl_gene_cleaned) <- colnames(methyl_gene_cleaned)[-1]

# Process clinical data (sort rownames)
sorted_clinical_focused_filtered_for_age <- clinical_focused_filtered_for_age[order(rownames(clinical_focused_filtered_for_age)), , drop = FALSE]
row_name_clinical <- rownames(sorted_clinical_focused_filtered_for_age)


# Merge them

# Check if the row names are consistent
all_row_names_match <- all(rownames(transposed_methyl_gene_cleaned) == row_name_clinical)
if (all_row_names_match) {
  gene_with_clinical_data <- cbind(sorted_clinical_focused_filtered_for_age, transposed_methyl_gene_cleaned)
  print("The row names are consistent and the data has been merged successfully.")
} else {
  print("The row names do not match. Please check the data.")
}
```



Now we perform Mann-Whitney test to compare HPV(+) and HPV(-) groups for each gene.



```{r Perform Mann-Whitney test and filter out significant results}
# Identify indices for HPV positive and negative patients
hpv_positive_indices <- which(gene_with_clinical_data$hpv_status == "positive")
hpv_negative_indices <- which(gene_with_clinical_data$hpv_status == "negative")

# Function to perform Mann-Whitney U test for a single gene
perform_test <- function(gene_index) {
  gene_values_positive <- gene_with_clinical_data[hpv_positive_indices, gene_index]
  gene_values_negative <- gene_with_clinical_data[hpv_negative_indices, gene_index]
  
  test_result <- wilcox.test(gene_values_positive, gene_values_negative)
  
  return(c(
    gene = colnames(gene_with_clinical_data)[gene_index],
    p_value = test_result$p.value,
    W = test_result$statistic
  ))
}

# Use parallel processing to speed up the computation
cl <- makeCluster(detectCores() - 1)
clusterExport(cl, c("gene_with_clinical_data", "hpv_positive_indices", "hpv_negative_indices", "perform_test"))

# Perform the test for all gene columns (10th to last)
test_results <- parLapply(cl, 10:ncol(gene_with_clinical_data), perform_test)
stopCluster(cl)

# Combine results into a data frame
test_results_df <- as.data.frame(do.call(rbind, test_results))
test_results_df$p_value <- as.numeric(as.character(test_results_df$p_value))
test_results_df$W <- as.numeric(as.character(test_results_df$W))

# Remove the duplicated W.W column
test_results_df$W.W <- NULL

# Format p-values to a specified number of significant digits
test_results_df$p_value <- format(test_results_df$p_value, digits = 4, scientific = TRUE)

# Save the results to a CSV file for easy checking
write.csv(test_results_df, "gene_comparison_results.csv", row.names = FALSE)

# Print a summary of significant results (e.g., p < 0.05)
significant_results <- test_results_df %>% filter(as.numeric(p_value) < 0.05)
print(significant_results)

# Save the results to a CSV file for easy checking
write.csv(significant_results, "gene_comparison_significant_results.csv", row.names = FALSE)
```




Now we roll back to the original gene data, filter out genes by HPV groups. This requires to extract two lists of patients' ID. We will still do it in Python.



```{r The gene(methylation) data that only has unique genes}
methyl_gene <- open_dataset("methyl_summed_genes.parquet")
methyl_gene <- as.data.frame(methyl_gene)
```


```{r}
clinical_all_CDEs <- fread("clinical_All_CDEs.txt", sep = "\t", header = TRUE)

# Transpose the clinical data
clinical_transposed <- t(clinical_all_CDEs)

clinical_transposed <- as.data.frame(clinical_transposed)

colnames(clinical_transposed) <- clinical_transposed[1,]

clinical_transposed <- clinical_transposed[-1,]

# Select necessary variables
clinical_focused_for_age <- clinical_transposed[, c(9, 12, 21:23, 46, 48, 54, 114)]

# Filter data
clinical_focused_filtered_for_age <- clinical_focused_for_age %>%
  filter(hpv_status != "indeterminate", tumor_tissue_site == "head and neck")
```



```{r Divide HPV groups and tract patients' ID}
# HPV+ group
hpv_positive <- clinical_focused_filtered_for_age %>% filter(hpv_status == "positive")

# HPV- group
hpv_negative <- clinical_focused_filtered_for_age %>% filter(hpv_status == "negative")


hpv_pos_patient <- rownames(hpv_positive)
hpv_neg_patient <- rownames(hpv_negative)

# Find the maximum length
max_length <- max(length(hpv_pos_patient), length(hpv_neg_patient))

# Pad the shorter list with NA values
hpv_pos_patient <- c(hpv_pos_patient, rep(NA, max_length - length(hpv_pos_patient)))
hpv_neg_patient <- c(hpv_neg_patient, rep(NA, max_length - length(hpv_neg_patient)))

# Create a data frame with the two lists
patient_lists <- data.frame(HPV_Positive = hpv_pos_patient, HPV_Negative = hpv_neg_patient)

# Write the data frame to a CSV file
write.csv(patient_lists, "patient_lists.csv", row.names = FALSE)
```




The 10th percentile value: 0.5131194491438932. We use 15% patient threshold (84 for positive group, 363 for negative group). 



```{python}
# import pandas as pd
# import pyarrow as pa
# import pyarrow.parquet as pq
# import numpy as np
# 
# # Load the Parquet file
# input_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_summed_genes.parquet"
# df = pd.read_parquet(input_file)
# 
# # Load the patient lists
# patient_lists = pd.read_csv("/Users/yufeiwang/Desktop/HNSCC/patient_lists.csv")
# hpv_pos_patient = patient_lists['HPV_Positive'].dropna().tolist()
# hpv_neg_patient = patient_lists['HPV_Negative'].dropna().tolist()
# 
# # Calculate the 10th percentile among all beta-values (excluding the first column)
# beta_values = df.iloc[:, 1:].values.flatten()
# percentile_10th_value = np.percentile(beta_values, 10)
# print(f"10th percentile value: {percentile_10th_value}")
# 
# # Filter out genes where >= 84 positive patients' beta-values are less than the 10th percentile value
# # AND >= 363 negative patients' beta-values are less than the 10th percentile value
# def filter_genes_by_group(row, pos_patients, neg_patients, threshold, pos_min, neg_min):
#     pos_values = row[pos_patients].astype(float)
#     neg_values = row[neg_patients].astype(float)
#     pos_condition = (pos_values < threshold).sum() >= pos_min
#     neg_condition = (neg_values < threshold).sum() >= neg_min
#     return not (pos_condition and neg_condition)
# 
# filtered_df = df[df.apply(lambda row: filter_genes_by_group(row, hpv_pos_patient, hpv_neg_patient, percentile_10th_value, 84, 363), axis=1)]
# 
# # Save the resulting dataset to a new Parquet file
# output_file = "/Users/yufeiwang/Desktop/HNSCC/methyl_cleaned_bygroup_genes.parquet"
# table = pa.Table.from_pandas(filtered_df)
# pq.write_table(table, output_file)
# 
# print(f"Filtered data saved to {output_file}")

```





```{r The data after cleaning genes by group}
methyl_gene_cleaned_bygroup <- open_dataset("methyl_cleaned_bygroup_genes.parquet")
methyl_gene_cleaned_bygroup <- as.data.frame(methyl_gene_cleaned_bygroup)

methyl_gene_cleaned_bygroup <- methyl_gene_cleaned_bygroup[, -ncol(methyl_gene_cleaned_bygroup)]
```



After filtering by group, there are 22859 genes left. Now we perform log2 transformation.



```{r Perform log transformation}
methyl_gene_cleaned_bygroup[, 2:ncol(methyl_gene_cleaned_bygroup)] <- log2(1 + methyl_gene_cleaned_bygroup[, 2:ncol(methyl_gene_cleaned_bygroup)])
```



Now we merge the clinical data with the new methylation (gene) data.




```{r Transpose the data to merge into the clinical data}
transposed_methyl_gene_cleaned_bygroup <- t(methyl_gene_cleaned_bygroup[, -1])

# Convert it to a data frame
transposed_methyl_gene_cleaned_bygroup <- as.data.frame(transposed_methyl_gene_cleaned_bygroup)

# Set the column names to the values in the first row (Gene Symbols)
colnames(transposed_methyl_gene_cleaned_bygroup) <- methyl_gene_cleaned_bygroup$Gene_Symbol

# Set the row names to the original column names (Patient IDs)
rownames(transposed_methyl_gene_cleaned_bygroup) <- colnames(methyl_gene_cleaned_bygroup)[-1]

# Process clinical data (sort rownames)
sorted_clinical_focused_filtered_for_age <- clinical_focused_filtered_for_age[order(rownames(clinical_focused_filtered_for_age)), , drop = FALSE]
row_name_clinical <- rownames(sorted_clinical_focused_filtered_for_age)


# Merge them

# Check if the row names are consistent
all_row_names_match <- all(rownames(transposed_methyl_gene_cleaned_bygroup) == row_name_clinical)
if (all_row_names_match) {
  gene_with_clinical_data_bygroup <- cbind(sorted_clinical_focused_filtered_for_age, transposed_methyl_gene_cleaned_bygroup)
  print("The row names are consistent and the data has been merged successfully.")
} else {
  print("The row names do not match. Please check the data.")
}
```




Now we perform Mann-Whitney test and save the test. (Add difference, FDR column; sort by difference: log2(Ratio)).



```{r perform Mann-Whitney test}
# Identify indices for HPV positive and negative patients
hpv_positive_indices <- which(gene_with_clinical_data_bygroup$hpv_status == "positive")
hpv_negative_indices <- which(gene_with_clinical_data_bygroup$hpv_status == "negative")

# Function to perform Mann-Whitney U test for a single gene
perform_test_bygroup <- function(gene_index) {
  gene_values_positive <- gene_with_clinical_data_bygroup[hpv_positive_indices, gene_index]
  gene_values_negative <- gene_with_clinical_data_bygroup[hpv_negative_indices, gene_index]
  
  # Only perform the test if both groups have enough non-zero values
  if (length(gene_values_positive[gene_values_positive != 0]) >= 2 & length(gene_values_negative[gene_values_negative != 0]) >= 2) {
    test_result <- wilcox.test(gene_values_positive, gene_values_negative)
    difference <- median(gene_values_positive) - median(gene_values_negative)
    ratio <- 2^difference - 1
    return(c(
      gene = colnames(gene_with_clinical_data_bygroup)[gene_index],
      p_value = test_result$p.value,
      difference = difference,
      ratio = ratio
    ))
  } else {
    return(c(
      gene = colnames(gene_with_clinical_data_bygroup)[gene_index],
      p_value = NA,
      difference = NA,
      ratio = NA
    ))
  }
}

# Use parallel processing to speed up the computation
cl <- makeCluster(detectCores() - 1)
clusterExport(cl, c("gene_with_clinical_data_bygroup", "hpv_positive_indices", "hpv_negative_indices", "perform_test"))

# Perform the test for all gene columns (10th to last)
test_results_bygroup <- parLapply(cl, 10:ncol(gene_with_clinical_data_bygroup), perform_test_bygroup)
stopCluster(cl)

# Combine results into a data frame
test_results_df_bygroup <- as.data.frame(do.call(rbind, test_results_bygroup))
test_results_df_bygroup$p_value <- as.numeric(as.character(test_results_df_bygroup$p_value))
test_results_df_bygroup$difference <- as.numeric(as.character(test_results_df_bygroup$difference))
test_results_df_bygroup$ratio <- as.numeric(as.character(test_results_df_bygroup$ratio))

# Calculate FDR
test_results_df_bygroup <- test_results_df_bygroup %>% 
  mutate(FDR = p.adjust(p_value, method = "fdr"))

# Format p-values to a specified number of significant digits
test_results_df_bygroup$p_value <- format(test_results_df_bygroup$p_value, digits = 4, scientific = TRUE)

# Sort the data frame by the difference column
test_results_df_bygroup <- test_results_df_bygroup %>% arrange(desc(difference))

# Save the results to a CSV file for easy checking
write.csv(test_results_df_bygroup, "gene_comparison_results_bygroup.csv", row.names = FALSE)

test_results_df_bygroup$p_value <- as.numeric(as.character(test_results_df_bygroup$p_value))
test_results_df_bygroup$FDR <- as.numeric(as.character(test_results_df_bygroup$FDR))
test_results_df_bygroup$difference <- as.numeric(as.character(test_results_df_bygroup$difference))

# Print the results
print(test_results_df_bygroup)

```




Now we keep genes with 1. p-value < 0.05 and 2. FDR < 0.001 to make volcano plots.




```{r Filter out genes such that p-value < 0.05 and make volcano plot} 
# Filter genes based on p-value < 0.05
filtered_data_pvalue <- test_results_df_bygroup %>%
  filter(as.numeric(p_value) < 0.05) %>%
  mutate(
    negLog10PValue = -log10(as.numeric(p_value)),
    significant = as.numeric(p_value) < 0.05
  )

# Create the volcano plot for p-value < 0.05
volcano_plot_pvalue <- ggplot(filtered_data_pvalue, aes(x = difference, y = negLog10PValue)) +
  geom_point(aes(color = significant)) +
  scale_color_manual(values = c("black", "red")) +
  labs(
    title = "Volcano Plot (p-value < 0.05)",
    x = "Log2 Fold Change",
    y = "-Log10 P-Value"
  ) +
  theme_minimal()

# Print the first volcano plot
print(volcano_plot_pvalue)



ggplot(filtered_data_pvalue, aes(x = difference, y = -log10(p_value))) +
  geom_point(aes(color = p_value < 0.05)) +
  scale_color_manual(values = c("black", "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  labs(title = "Volcano Plot (p-value < 0.05)", x = "Log2 Fold Change", y = "-Log10 P-Value") +
  theme_minimal()

```








```{r Filter out genes such that FDR < 0.001 and make volcano plot}
# Filter genes based on FDR < 0.001
filtered_data_fdr <- test_results_df_bygroup %>%
  filter(as.numeric(FDR) < 0.001) %>%
  mutate(
    negLog10PValue = -log10(as.numeric(p_value)),
    significant = as.numeric(FDR) < 0.001
  )

# Create the volcano plot for FDR < 0.001
volcano_plot_fdr <- ggplot(filtered_data_fdr, aes(x = difference, y = negLog10PValue)) +
  geom_point(aes(color = significant)) +
  scale_color_manual(values = c("black", "red")) +
  labs(
    title = "Volcano Plot (FDR < 0.001)",
    x = "Log2 Fold Change",
    y = "-Log10 P-Value"
  ) +
  theme_minimal()

# Print the second volcano plot
print(volcano_plot_fdr)



# Create the volcano plot for the FDR filtered data
ggplot(filtered_data_fdr, aes(x = difference, y = -log10(p_value))) +
  geom_point(aes(color = FDR < 0.001)) +
  scale_color_manual(values = c("black", "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  labs(title = "Volcano Plot (FDR < 0.001)", x = "Log2 Fold Change", y = "-Log10 P-Value") +
  theme_minimal()


```





Let's roll back to the testing part, now we use t-test for all genes. And create some more columns: log2(ratio), this is the same as difference; ratio; FC (ratio or $-\frac{1}{ratio}$ depending on it's greater than or less than 1); p-value; FDR.





```{r perform t-test}
# Identify indices for HPV positive and negative patients
hpv_positive_indices <- which(gene_with_clinical_data_bygroup$hpv_status == "positive")
hpv_negative_indices <- which(gene_with_clinical_data_bygroup$hpv_status == "negative")

# Function to perform t-test for a single gene
perform_test_bygroup_ttest <- function(gene_index) {
  gene_values_positive <- gene_with_clinical_data_bygroup[hpv_positive_indices, gene_index]
  gene_values_negative <- gene_with_clinical_data_bygroup[hpv_negative_indices, gene_index]
  
  # Only perform the test if both groups have enough non-zero values
  if (length(gene_values_positive[gene_values_positive != 0]) >= 2 & length(gene_values_negative[gene_values_negative != 0]) >= 2) {
    test_result <- t.test(gene_values_positive, gene_values_negative)
    difference <- test_result$estimate[1] - test_result$estimate[2]
    ratio <- 2^difference
    FC <- ifelse(ratio > 1, ratio, -1/ratio)
    
    return(c(
      gene = colnames(gene_with_clinical_data_bygroup)[gene_index],
      `difference_log2_scale (pos - neg)` = difference,
      `ratio (pos / neg)` = ratio,
      FC = FC,
      p_value = test_result$p.value
    ))
  } else {
    return(c(
      gene = colnames(gene_with_clinical_data_bygroup)[gene_index],
      `difference_log2_scale (pos - neg)` = NA,
      `ratio (pos / neg)` = NA,
      FC = NA,
      p_value = NA
    ))
  }
}

# Use parallel processing to speed up the computation
cl <- makeCluster(detectCores() - 1)
clusterExport(cl, c("gene_with_clinical_data_bygroup", "hpv_positive_indices", "hpv_negative_indices", "perform_test_bygroup_ttest"))

# Perform the test for all gene columns (10th to last)
test_results_bygroup_ttest <- parLapply(cl, 10:ncol(gene_with_clinical_data_bygroup), perform_test_bygroup_ttest)
stopCluster(cl)

# Combine results into a data frame
test_results_df_bygroup_ttest <- as.data.frame(do.call(rbind, test_results_bygroup_ttest))
test_results_df_bygroup_ttest$p_value <- as.numeric(as.character(test_results_df_bygroup_ttest$p_value))
test_results_df_bygroup_ttest$`difference_log2_scale (pos - neg)` <- as.numeric(as.character(test_results_df_bygroup_ttest$`difference_log2_scale (pos - neg)`))
test_results_df_bygroup_ttest$`ratio (pos / neg)` <- as.numeric(as.character(test_results_df_bygroup_ttest$`ratio (pos / neg)`))
test_results_df_bygroup_ttest$FC <- as.numeric(as.character(test_results_df_bygroup_ttest$FC))

# Calculate FDR
test_results_df_bygroup_ttest <- test_results_df_bygroup_ttest %>% 
  mutate(FDR = p.adjust(p_value, method = "fdr"))

# Sort the data frame by the difference column
test_results_df_bygroup_ttest <- test_results_df_bygroup_ttest %>% arrange(desc(`difference_log2_scale (pos - neg)`))

# Remove 2nd to 4th columns that are character types
test_results_df_bygroup_ttest <- test_results_df_bygroup_ttest[, -c(2, 3, 4)]

# Save the results to a CSV file for easy checking
write.csv(test_results_df_bygroup_ttest, "gene_comparison_results_bygroup_ttest.csv", row.names = FALSE)

# Print the results
print(test_results_df_bygroup_ttest)

```


Make volcano plots using all genes.



```{r Plot based on difference_log2_scale (pos - neg)}
# Determine genes outside of the vertical lines
significant_genes_difference <- test_results_df_bygroup_ttest %>%
  filter(`difference_log2_scale (pos - neg)` > log2(1.3) | `difference_log2_scale (pos - neg)` < -log2(1.3))

# Select the top 10 largest and lowest genes based on difference_log2_scale (pos - neg)
top_genes_diff <- significant_genes_difference %>%
  arrange(desc(`difference_log2_scale (pos - neg)`)) %>%
  head(10)

lowest_genes_diff <- significant_genes_difference %>%
  arrange(`difference_log2_scale (pos - neg)`) %>%
  head(10)

# Combine top and lowest genes
label_genes_diff <- bind_rows(top_genes_diff, lowest_genes_diff)

# Create the volcano plot
volcano_plot_difference <- ggplot(test_results_df_bygroup_ttest, aes(x = `difference_log2_scale (pos - neg)`, y = -log10(p_value))) +
  geom_point(aes(color = p_value < 0.05), alpha = 0.8, size = 1.5) +
  scale_color_manual(values = c("black", "red")) +
  geom_vline(xintercept = c(-0.58, 0.58), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_text_repel(data = label_genes_diff, aes(label = gene), size = 3) +
  labs(title = "Volcano Plot (T-test Results)", x = "Difference_log2_scale (pos - neg)", y = "-Log10 P-Value") +
  theme_minimal()

# Print the plot
print(volcano_plot_difference)
```





```{r Plot based on ratio (pos/neg)}
# Identify significant genes based on ratio
significant_genes_ratio <- test_results_df_bygroup_ttest %>%
  filter(`ratio (pos / neg)` > 1.3 | `ratio (pos / neg)` < 1/1.3)

# Select the top 10 largest and lowest genes based on ratio
top_genes_ratio <- significant_genes_ratio %>%
  arrange(desc(`ratio (pos / neg)`)) %>%
  head(10)

lowest_genes_ratio <- significant_genes_ratio %>%
  arrange(`ratio (pos / neg)`) %>%
  head(10)

# Combine top and lowest genes
label_genes_ratio <- bind_rows(top_genes_ratio, lowest_genes_ratio)

# Create volcano plot based on ratio (title = "Volcano Plot (T-test Results)")
volcano_plot_ratio <- ggplot(test_results_df_bygroup_ttest, aes(x = `ratio (pos / neg)`, y = -log10(p_value))) +
  geom_point(aes(color = p_value < 0.05), alpha = 0.8, size = 1.5) +
  scale_color_manual(values = c("black", "red")) +
  geom_vline(xintercept = c(1/1.3, 1.3), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_text_repel(data = label_genes_ratio, aes(label = gene), size = 3) +
  labs(x = "Ratio (pos / neg)", y = "-Log10 P-Value") +
  theme_minimal()

# Save the plot as a PNG file with 300 DPI
ggsave("Figure_2A_colcano.png", plot = volcano_plot_ratio, width = 10, height = 8, units = "in", dpi = 300)

# Print the plot
print(volcano_plot_ratio)
```



$$NOTE:\ Update\ my\ volcano\ plot$$


```{r Update my colcano plot}
test_results_df_bygroup_ttest_copy <- test_results_df_bygroup_ttest

# Update the significant_genes_ratio data frame with new labels
test_results_df_bygroup_ttest_copy$Methylation <- "NO"
test_results_df_bygroup_ttest_copy$Methylation[test_results_df_bygroup_ttest_copy$`ratio (pos / neg)` > 1.3] <- "UP"
test_results_df_bygroup_ttest_copy$Methylation[test_results_df_bygroup_ttest_copy$`ratio (pos / neg)` < 1/1.3] <- "DOWN"

# Set colors for the new labels
mycolors <- c("green", "red", "grey")
names(mycolors) <- c("DOWN", "UP", "NO")

# Identify significant genes based on ratio
significant_genes_ratio <- test_results_df_bygroup_ttest_copy %>%
  filter(`ratio (pos / neg)` > 1.3 | `ratio (pos / neg)` < 1/1.3)

# Select the top 10 largest and lowest genes based on ratio
top_genes_ratio <- significant_genes_ratio %>%
  arrange(desc(`ratio (pos / neg)`)) %>%
  head(10)

lowest_genes_ratio <- significant_genes_ratio %>%
  arrange(`ratio (pos / neg)`) %>%
  head(10)

# Combine top and lowest genes
label_genes_ratio <- bind_rows(top_genes_ratio, lowest_genes_ratio)

# Update the delabel column for the genes to be labeled
test_results_df_bygroup_ttest_copy$delabel <- NA
test_results_df_bygroup_ttest_copy$delabel[test_results_df_bygroup_ttest_copy$Methylation != "NO"] <- 
  test_results_df_bygroup_ttest_copy$gene[test_results_df_bygroup_ttest_copy$Methylation != "NO"]

# Create the volcano plot with updated labeling and coloring
volcano_plot_ratio <- ggplot(test_results_df_bygroup_ttest_copy, aes(x = `ratio (pos / neg)`, y = -log10(p_value), col = Methylation)) +
  geom_point(alpha = 0.8, size = 1.5) +
  geom_vline(xintercept = c(1/1.3, 1.3), col = "blue", linetype = "dashed") +
  geom_hline(yintercept = 1.3, col = "blue", linetype = "dashed") +
  geom_text_repel(data = label_genes_ratio, aes(label = gene), size = 3, fontface = "bold", show.legend = FALSE) +
  scale_color_manual(values = mycolors, breaks = c("DOWN", "UP", "NO")) +
  labs(x = "Ratio (pos / neg)", y = "-Log10 P-Value", color = "Methylation") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), plot.title = element_text(hjust = 0.5)) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

# Save the plot as a PNG file with 300 DPI
ggsave("Figure_2A_volcano_new.png", plot = volcano_plot_ratio, width = 10, height = 8, units = "in", dpi = 300)

# Print the plot
print(volcano_plot_ratio)

```




$$END\ NOTE$$



```{r Plot based on FC}
# Identify significant genes based on FC
significant_genes_fc <- test_results_df_bygroup_ttest %>%
  filter(FC < -1.3 | FC > 1.3)

# Select the top 10 largest and lowest genes based on FC
top_genes_fc <- significant_genes_fc %>%
  arrange(desc(FC)) %>%
  head(10)

lowest_genes_fc <- significant_genes_fc %>%
  arrange(FC) %>%
  head(10)

# Combine top and lowest genes
label_genes_fc <- bind_rows(top_genes_fc, lowest_genes_fc)

# Create volcano plot based on FC
volcano_plot_fc <- ggplot(test_results_df_bygroup_ttest, aes(x = FC, y = -log10(p_value))) +
  geom_point(aes(color = p_value < 0.05), alpha = 0.8, size = 1.5) +
  scale_color_manual(values = c("black", "red")) +
  geom_vline(xintercept = c(-1.3, 1.3), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_text_repel(data = label_genes_fc, aes(label = gene), size = 3) +
  labs(title = "Volcano Plot (T-test Results)", x = "Fold Change (FC)", y = "-Log10 P-Value") +
  theme_minimal()

print(volcano_plot_fc)
```




```{r Plot based on FDR}
# Determine genes outside of the vertical lines based on FDR
significant_genes_fdr <- test_results_df_bygroup_ttest %>%
  filter(FDR < 0.001)

# Create the volcano plot based on FDR
volcano_plot_fdr <- ggplot(test_results_df_bygroup_ttest, aes(x = `difference_log2_scale (pos - neg)`, y = -log10(FDR))) +
  geom_point(aes(color = FDR < 0.001), alpha = 0.8, size = 1.5) +
  scale_color_manual(values = c("black", "red")) +
  geom_vline(xintercept = c(-0.58, 0.58), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "blue") +
  labs(title = "Volcano Plot (FDR < 0.001)", x = "Difference_log2_scale (pos - neg)", y = "-Log10 FDR") +
  theme_minimal()

# Print the plot
print(volcano_plot_fdr)

```


We will not create more volcano plots for FDR based on other scales. We now will create the table that has all genes (22859), beta-values (no log transformation), t-test results. After this we will create heatmap based on the filtered genes (93).



```{r This is the data with beta-value in original scale for all unique genes and for all patients}
methyl_gene_cleaned_bygroup_ori <- open_dataset("methyl_cleaned_bygroup_genes.parquet")
methyl_gene_cleaned_bygroup_ori <- as.data.frame(methyl_gene_cleaned_bygroup_ori)

methyl_gene_cleaned_bygroup_ori <- methyl_gene_cleaned_bygroup_ori[, -ncol(methyl_gene_cleaned_bygroup_ori)]
```



```{r Merge t-test result data to the beta-value data}
# Assuming `test_results_df_bygroup_ttest` is the test result data frame and `beta_value_data` is the beta-value data frame
test_results_df_bygroup_ttest <- as.data.frame(test_results_df_bygroup_ttest)
methyl_gene_cleaned_bygroup_ori <- as.data.frame(methyl_gene_cleaned_bygroup_ori)

# Set row names as gene names
rownames(test_results_df_bygroup_ttest) <- test_results_df_bygroup_ttest$gene
rownames(methyl_gene_cleaned_bygroup_ori) <- methyl_gene_cleaned_bygroup_ori$Gene_Symbol

# Remove the gene columns after setting them as row names
test_results_df_bygroup_ttest <- test_results_df_bygroup_ttest[, -1]
methyl_gene_cleaned_bygroup_ori <- methyl_gene_cleaned_bygroup_ori[, -1]

# Sort both data frames by row names (gene names)
test_results_df_bygroup_ttest <- test_results_df_bygroup_ttest[order(rownames(test_results_df_bygroup_ttest)), ]
methyl_gene_cleaned_bygroup_ori <- methyl_gene_cleaned_bygroup_ori[order(rownames(methyl_gene_cleaned_bygroup_ori)), ]

# Check if all row names are the same and in the same order
rownames_match <- all(rownames(test_results_df_bygroup_ttest) == rownames(methyl_gene_cleaned_bygroup_ori))

# If row names match, merge the data frames
if (rownames_match) {
  # Merge the data frames by row names
  merged_beta_ttest_table <- cbind(test_results_df_bygroup_ttest, methyl_gene_cleaned_bygroup_ori)
  print("The data frames have been merged successfully.")
} else {
  print("The row names do not match.")
}

# Sort the merged data by the `difference` column for clear heatmap visualization
merged_beta_ttest_table <- merged_beta_ttest_table[order(merged_beta_ttest_table$`difference_log2_scale (pos - neg)`, decreasing = TRUE), ]

# Save the merged data to a CSV file
write.csv(merged_beta_ttest_table, "merged_beta_ttest_table.csv", row.names = TRUE)
```



The filtered data is based on the ratio < 1/1.3, ratio > 1.3, and p-value < 0.05 thresholds.




```{r Filter out genes in the table}
# Get the list of significant gene names
significant_gene_names <- significant_genes_ratio$gene

# Filter the merged data to keep only the significant genes
merged_beta_ttest_table_filtered <- merged_beta_ttest_table[rownames(merged_beta_ttest_table) %in% significant_gene_names, ]

# Ensure all p-values are less than 0.05
merged_beta_ttest_table_filtered <- merged_beta_ttest_table_filtered[merged_beta_ttest_table_filtered$p_value < 0.05, ]

write.csv(merged_beta_ttest_table_filtered, "merged_beta_ttest_table_filtered.csv", row.names = TRUE)
```




$$ NOTE: Section\ below\ is\ for\ the\ poster$$

```{r Generate table 2 (93 genes table)}
table2 <- merged_beta_ttest_table_filtered[, c(1,4,5)]

table2$Gene_Symbol <- rownames(table2)
rownames(table2) <- NULL

# Rearrange columns to have Gene_Symbol as the first column
table2 <- table2[, c("Gene_Symbol", "p_value", "FC", "FDR")]

# Adjust the columns' formats
# table2$p_value <- formatC(table2$p_value, format = "e", digits = 2)
# table2$FDR <- formatC(table2$FDR, format = "e", digits = 2)
# table2$FC <- formatC(table2$FC, format = "f", digits = 2)

write_xlsx(table2, "93_genes_table.xlsx")

largest_FC <- table2[order(-table2$FC), ][1:10, ]
smallest_FC <- table2[order(table2$FC), ][1:10, ]

# Flip the smallest_FC to be in ascending order
smallest_FC <- smallest_FC[order(-smallest_FC$FC), ]

# Adjust the columns' formats
largest_FC$p_value <- formatC(largest_FC$p_value, format = "e", digits = 2)
largest_FC$FDR <- formatC(largest_FC$FDR, format = "e", digits = 2)
largest_FC$FC <- formatC(largest_FC$FC, format = "f", digits = 2)

smallest_FC$p_value <- formatC(smallest_FC$p_value, format = "e", digits = 2)
smallest_FC$FDR <- formatC(smallest_FC$FDR, format = "e", digits = 2)
smallest_FC$FC <- formatC(smallest_FC$FC, format = "f", digits = 2)

# Combine the two subsets and add '...' in the middle
middle_row <- data.frame(Gene_Symbol = '...', p_value = '...', FC = '...', FDR = '...')
table2_new <- rbind(largest_FC, middle_row, smallest_FC)

table_grob <- tableGrob(table2_new, rows = NULL)

# Save the table as a PNG file
png(filename = "table2.png", width = 10, height = 12, units = "in", res = 100)
grid.draw(table_grob)
dev.off()

```



$$END\ NOTE$$



Now group the patients by HPV status (positive first) for both data.



```{r Group patients for original data based on hpv_status (positive first)}
# HPV+ group
hpv_positive <- clinical_focused_filtered_for_age %>% filter(hpv_status == "positive")

# HPV- group
hpv_negative <- clinical_focused_filtered_for_age %>% filter(hpv_status == "negative")


hpv_pos_patient <- rownames(hpv_positive)
hpv_neg_patient <- rownames(hpv_negative)

# Extract the patient columns from the merged data
patient_columns <- colnames(merged_beta_ttest_table)[6:ncol(merged_beta_ttest_table)]

# Verify if all patient columns are accounted for
all_match <- all(patient_columns %in% c(hpv_pos_patient, hpv_neg_patient))

if (all_match) {
  # Rearrange the columns: positive patients first, followed by negative patients
  ordered_patient_columns <- c(hpv_pos_patient, hpv_neg_patient)
  merged_beta_ttest_table_grouped <- merged_beta_ttest_table %>%
    dplyr::select(p_value, `difference_log2_scale (pos - neg)`, `ratio (pos / neg)`, FC, FDR, all_of(ordered_patient_columns))
  print("Patients are grouped.")
} else {
  print("Patient ID not match.")
}

write.csv(merged_beta_ttest_table_grouped, "merged_beta_ttest_table.csv", row.names = TRUE)
```




```{r Group pattients for filtered data based on hpv_status (positive first)}
# Extract the patient columns from the merged data
patient_columns <- colnames(merged_beta_ttest_table_filtered)[6:ncol(merged_beta_ttest_table_filtered)]

# Verify if all patient columns are accounted for
all_match <- all(patient_columns %in% c(hpv_pos_patient, hpv_neg_patient))

if (all_match) {
  # Rearrange the columns: positive patients first, followed by negative patients
  ordered_patient_columns <- c(hpv_pos_patient, hpv_neg_patient)
  merged_beta_ttest_table_filtered_grouped <- merged_beta_ttest_table_filtered %>%
    dplyr::select(p_value, `difference_log2_scale (pos - neg)`, `ratio (pos / neg)`, FC, FDR, all_of(ordered_patient_columns))
  print("Patients are grouped.")
} else {
  print("Patient ID not match.")
}

write.csv(merged_beta_ttest_table_filtered_grouped, "merged_beta_ttest_table_filtered.csv", row.names = TRUE)
```



Add HPV status row for both data.



```{r Add HPV_status for original data}
# Create a row indicating HPV status for each patient
hpv_status_row <- c("NA", "NA", "NA", "NA", "NA", 
                    ifelse(colnames(merged_beta_ttest_table_grouped)[6:ncol(merged_beta_ttest_table_grouped)] %in% hpv_pos_patient, "positive", "negative"))

# Combine this row with the existing data frame
merged_beta_ttest_table_grouped_with_status <- rbind(hpv_status_row, merged_beta_ttest_table_grouped)

# Rename the columns to avoid conflict
colnames(merged_beta_ttest_table_grouped_with_status) <- c("p_value", "difference_log2_scale (pos - neg)", "ratio (pos / neg)", "FC", "FDR",
                                                                    colnames(merged_beta_ttest_table_grouped)[6:ncol(merged_beta_ttest_table_grouped)])

rownames(merged_beta_ttest_table_grouped_with_status)[1] <- "HPV_status"

write.csv(merged_beta_ttest_table_grouped_with_status, "merged_beta_ttest_table_grouped_with_status.csv", row.names = TRUE)
```



```{r Add HPV_status for filtered data}
# Create a row indicating HPV status for each patient
hpv_status_row <- c("NA", "NA", "NA", "NA", "NA", 
                    ifelse(colnames(merged_beta_ttest_table_filtered_grouped)[6:ncol(merged_beta_ttest_table_filtered_grouped)] %in% hpv_pos_patient, "positive", "negative"))

# Combine this row with the existing data frame
merged_beta_ttest_table_filtered_grouped_with_status <- rbind(hpv_status_row, merged_beta_ttest_table_filtered_grouped)

# Rename the columns to avoid conflict
colnames(merged_beta_ttest_table_filtered_grouped_with_status) <- c("p_value", "difference_log2_scale (pos - neg)", "ratio (pos / neg)", "FC", "FDR",
                                                                    colnames(merged_beta_ttest_table_filtered_grouped)[6:ncol(merged_beta_ttest_table_filtered_grouped)])

rownames(merged_beta_ttest_table_filtered_grouped_with_status)[1] <- "HPV_status"

write.csv(merged_beta_ttest_table_filtered_grouped_with_status, "merged_beta_ttest_table_filtered_grouped_with_status.csv", row.names = TRUE)
```



Next we will create the heatmap.



```{r Heatmap}
# Convert the data to numeric format for heatmap plotting
heatmap_data <- as.matrix(merged_beta_ttest_table_filtered_grouped_with_status[-1, -(1:5)])  # Exclude the first row (HPV status) and the first 5 columns (metadata)
heatmap_data <- apply(heatmap_data, 2, as.numeric)

# Create a data frame for annotation
annotation_col <- data.frame(
  HPV_status = factor(ifelse(colnames(heatmap_data) %in% hpv_pos_patient, "positive", "negative"))
)
rownames(annotation_col) <- colnames(heatmap_data)

# Plot the heatmap
pheatmap(heatmap_data, 
         annotation_col = annotation_col, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE, 
         show_rownames = TRUE, 
         show_colnames = FALSE, 
         scale = "row", 
         color = colorRampPalette(c("blue", "white", "red"))(50),
         annotation_colors = list(HPV_status = c("positive" = "red", "negative" = "blue")),
         main = "Gene Methylation by HPV Status",
         fontsize_row = 6)

```





```{r Heatmap with label, and save as .png file}
# Set the row labels for the heatmap
row_labels <- rownames(merged_beta_ttest_table_filtered_grouped_with_status)[-1]

# Save the heatmap as a high-quality PNG file
png("high_quality_heatmap.png", width = 3600, height = 2400, res = 300)

# Plot the heatmap
pheatmap(heatmap_data, 
         annotation_col = annotation_col, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE, 
         show_rownames = TRUE, 
         show_colnames = FALSE, 
         scale = "row", 
         color = colorRampPalette(c("blue", "white", "red"))(50),
         annotation_colors = list(HPV_status = c("positive" = "red", "negative" = "blue")),
         #main = "Gene Methylation by HPV Status",
         fontsize_row = 8,  # Adjusted for smaller font size
         labels_row = row_labels  # Ensures that row labels are displayed
)

# Close the PNG device
dev.off()
```





At this point, we will start to create all plots for the poster. The poster includes:


### Table 1 (clinical table)

Figure 1A: box plot, methylation between HPV(+) and HPV(-)

Figure 1B: box plot, methylation Female HPV+, Female HPV (-), male HPV (+), male HPV (-)

Note: Label Y-axis as global methylation

 
### Table 2: List genes with FC of 1.3 with p-value<0.5, include FC and p-values, FDR

Figure 2A: Volcano plot

Figure 2B: Heatmap

 
### Figure 3: Pathway analysis




```{r Create the clinical table}
clinical_all_CDEs <- fread("clinical_All_CDEs.txt", sep = "\t", header = TRUE)

# Transpose the clinical data
clinical_transposed <- t(clinical_all_CDEs)

clinical_transposed <- as.data.frame(clinical_transposed)

colnames(clinical_transposed) <- clinical_transposed[1,]

clinical_transposed <- clinical_transposed[-1,]

# Select necessary variables
clinical_focused_for_age <- clinical_transposed[, c(9, 12, 21:23, 46, 48, 54, 114)]

# Filter data
clinical_focused_filtered_for_age <- clinical_focused_for_age %>%
  filter(hpv_status != "indeterminate", tumor_tissue_site == "head and neck")

clinical_focused_filtered_for_age$age_at_initial_pathologic_diagnosis <- as.numeric(clinical_focused_filtered_for_age$age_at_initial_pathologic_diagnosis)
```



```{r Table 1 version 1}
# Ensure the hpv_status values are consistent
clinical_focused_filtered_for_age$hpv_status <- ifelse(clinical_focused_filtered_for_age$hpv_status %in% c("HPV(+)", "positive"), "HPV(+)", 
                                                       ifelse(clinical_focused_filtered_for_age$hpv_status %in% c("HPV(-)", "negative"), "HPV(-)", NA))

# Calculate the summary statistics
clinical_summary <- clinical_focused_filtered_for_age %>%
  group_by(hpv_status) %>%
  summarise(
    Mean_Age = round(mean(age_at_initial_pathologic_diagnosis, na.rm = TRUE), 2),
    Median_Age = median(age_at_initial_pathologic_diagnosis, na.rm = TRUE),
    Range_Age = paste0(min(age_at_initial_pathologic_diagnosis, na.rm = TRUE), " - ", max(age_at_initial_pathologic_diagnosis, na.rm = TRUE)),
    Female_Count = sum(gender == "female", na.rm = TRUE),
    Male_Count = sum(gender == "male", na.rm = TRUE),
    Total_Count = n(),
    .groups = "drop"
  ) %>%
  mutate(
    Female_Percentage = round(Female_Count / Total_Count * 100, 2),
    Male_Percentage = round(Male_Count / Total_Count * 100, 2)
  )

# Extract subdivision counts as a data frame and calculate percentages
subdivision_counts <- clinical_focused_filtered_for_age %>%
  group_by(hpv_status, anatomic_neoplasm_subdivision) %>%
  summarise(Count = n(), .groups = "drop") %>%
  spread(hpv_status, Count, fill = 0) %>%
  mutate(
    Total = `HPV(+)` + `HPV(-)`,
    HPV_positive_percentage = round(`HPV(+)` / sum(`HPV(+)`, na.rm = TRUE) * 100, 2),
    HPV_negative_percentage = round(`HPV(-)` / sum(`HPV(-)`, na.rm = TRUE) * 100, 2)
  )

# Prepare the table content
table_content <- data.frame(
  Category = c("Age", "", "", "Gender (%)", "", "Subdivision (%)", rep("", nrow(subdivision_counts) - 1)),
  Subcategory = c("Mean", "Median", "Range", "Female", "Male", subdivision_counts$anatomic_neoplasm_subdivision),
  `HPV(+)` = c(clinical_summary$Mean_Age[clinical_summary$hpv_status == "HPV(+)"],
               clinical_summary$Median_Age[clinical_summary$hpv_status == "HPV(+)"],
               clinical_summary$Range_Age[clinical_summary$hpv_status == "HPV(+)"],
               paste0(clinical_summary$Female_Count[clinical_summary$hpv_status == "HPV(+)"], " (", clinical_summary$Female_Percentage[clinical_summary$hpv_status == "HPV(+)"], ")"),
               paste0(clinical_summary$Male_Count[clinical_summary$hpv_status == "HPV(+)"], " (", clinical_summary$Male_Percentage[clinical_summary$hpv_status == "HPV(+)"], ")"),
               paste0(as.integer(subdivision_counts$`HPV(+)`), " (", subdivision_counts$HPV_positive_percentage, ")")),
  `HPV(-)` = c(clinical_summary$Mean_Age[clinical_summary$hpv_status == "HPV(-)"],
               clinical_summary$Median_Age[clinical_summary$hpv_status == "HPV(-)"],
               clinical_summary$Range_Age[clinical_summary$hpv_status == "HPV(-)"],
               paste0(clinical_summary$Female_Count[clinical_summary$hpv_status == "HPV(-)"], " (", clinical_summary$Female_Percentage[clinical_summary$hpv_status == "HPV(-)"], ")"),
               paste0(clinical_summary$Male_Count[clinical_summary$hpv_status == "HPV(-)"], " (", clinical_summary$Male_Percentage[clinical_summary$hpv_status == "HPV(-)"], ")"),
               paste0(as.integer(subdivision_counts$`HPV(-)`), " (", subdivision_counts$HPV_negative_percentage, ")"))
)

# Display the table
table_content %>%
  kable("html", col.names = c("", "", "HPV(+)", "HPV(-)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```







```{r Table 1 version 2 (selected)}
# Ensure the hpv_status values are consistent
clinical_focused_filtered_for_age$hpv_status <- ifelse(clinical_focused_filtered_for_age$hpv_status %in% c("HPV(+)", "positive"), "HPV(+)", 
                                                       ifelse(clinical_focused_filtered_for_age$hpv_status %in% c("HPV(-)", "negative"), "HPV(-)", NA))

# Calculate the summary statistics
clinical_summary <- clinical_focused_filtered_for_age %>%
  group_by(hpv_status) %>%
  summarise(
    Mean_Age = round(mean(age_at_initial_pathologic_diagnosis, na.rm = TRUE), 2),
    Median_Age = round(median(age_at_initial_pathologic_diagnosis, na.rm = TRUE), 2),
    Range_Age = paste0(min(age_at_initial_pathologic_diagnosis, na.rm = TRUE), " - ", max(age_at_initial_pathologic_diagnosis, na.rm = TRUE)),
    Female_Count = sum(gender == "female", na.rm = TRUE),
    Male_Count = sum(gender == "male", na.rm = TRUE),
    Total_Count = n(),
    .groups = "drop"
  ) %>%
  mutate(
    Female_Percentage = round(Female_Count / Total_Count * 100, 2),
    Male_Percentage = round(Male_Count / Total_Count * 100, 2)
  )

# Extract subdivision counts as a data frame and calculate percentages
subdivision_counts <- clinical_focused_filtered_for_age %>%
  group_by(hpv_status, anatomic_neoplasm_subdivision) %>%
  summarise(Count = n(), .groups = "drop") %>%
  spread(hpv_status, Count, fill = 0) %>%
  mutate(
    Total = `HPV(+)` + `HPV(-)`,
    HPV_positive_percentage = round(`HPV(+)` / sum(`HPV(+)`, na.rm = TRUE) * 100, 2),
    HPV_negative_percentage = round(`HPV(-)` / sum(`HPV(-)`, na.rm = TRUE) * 100, 2)
  )

# Prepare the table content
table_content <- data.frame(
  Category = c("Age", "", "", "Sex", "", "Subdivision", rep("", nrow(subdivision_counts) - 1)),
  Subcategory = c("Mean", "Median", "Range", "Female", "Male", subdivision_counts$anatomic_neoplasm_subdivision),
  `HPV(+)` = c(clinical_summary$Mean_Age[clinical_summary$hpv_status == "HPV(+)"],
               formatC(clinical_summary$Median_Age[clinical_summary$hpv_status == "HPV(+)"], format = "f", digits = 2),
               clinical_summary$Range_Age[clinical_summary$hpv_status == "HPV(+)"],
               paste0(clinical_summary$Female_Count[clinical_summary$hpv_status == "HPV(+)"], " (", formatC(clinical_summary$Female_Percentage[clinical_summary$hpv_status == "HPV(+)"], format = "f", digits = 2), "%)"),
               paste0(clinical_summary$Male_Count[clinical_summary$hpv_status == "HPV(+)"], " (", formatC(clinical_summary$Male_Percentage[clinical_summary$hpv_status == "HPV(+)"], format = "f", digits = 2), "%)"),
               paste0(as.integer(subdivision_counts$`HPV(+)`), " (", formatC(subdivision_counts$HPV_positive_percentage, format = "f", digits = 2), "%)")),
  `HPV(-)` = c(clinical_summary$Mean_Age[clinical_summary$hpv_status == "HPV(-)"],
               formatC(clinical_summary$Median_Age[clinical_summary$hpv_status == "HPV(-)"], format = "f", digits = 2),
               clinical_summary$Range_Age[clinical_summary$hpv_status == "HPV(-)"],
               paste0(clinical_summary$Female_Count[clinical_summary$hpv_status == "HPV(-)"], " (", formatC(clinical_summary$Female_Percentage[clinical_summary$hpv_status == "HPV(-)"], format = "f", digits = 2), "%)"),
               paste0(clinical_summary$Male_Count[clinical_summary$hpv_status == "HPV(-)"], " (", formatC(clinical_summary$Male_Percentage[clinical_summary$hpv_status == "HPV(-)"], format = "f", digits = 2), "%)"),
               paste0(as.integer(subdivision_counts$`HPV(-)`), " (", formatC(subdivision_counts$HPV_negative_percentage, format = "f", digits = 2), "%)"))
)

# Display the table with improved formatting
html_table <- table_content %>%
  kable("html", col.names = c("", "", "HPV(+)", "HPV(-)"), align = c("l", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left") %>%
  column_spec(1, bold = TRUE, border_left = TRUE) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(3, border_right = TRUE) %>%
  column_spec(4, border_right = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria")

# Add CSS for vertical lines
html_table <- gsub('<table class="kable_table', '<table class="kable_table" style="border-collapse: collapse; border-left: 2px solid black; border-right: 2px solid black;', html_table)

html_table <- gsub('<tbody>', '<tbody style="border-left: 2px solid black; border-right: 2px solid black;">', html_table)
html_table <- gsub('<th style="', '<th style="border-left: 2px solid black; border-right: 2px solid black; ', html_table)
html_table <- gsub('<td style="', '<td style="border-left: 2px solid black; border-right: 2px solid black; ', html_table)

# Ensure no vertical border between Category and Subcategory
html_table <- gsub('<td style="border-right: 2px solid black; ', '<td style="', html_table, fixed = TRUE)

html_table

save_kable(html_table, "table1_new.html")

# Use webshot to save the HTML as a PNG with specified dpi
webshot("table1_new.html", "table1_new.png", zoom = 3, vwidth = 1000, vheight = 800)

img <- image_read("table1_new.png")

# Rescale the image to ensure it is 100 dpi
img <- image_scale(img, "1000x800")

# Save the image with 100 dpi
image_write(img, path = "table1_new_100dpi.png", density = 800)

# Read the image
img <- image_read("table1_new.png")

# Optionally, rescale the image (adjust the dimensions if necessary)
# For example, scaling to a desired size while maintaining proportions
img <- image_scale(img, "1500x1200")

# Save the image with the specified DPI
image_write(img, path = "table1_new_700dpi.png", density = 700)


```



I will do the 5-year survival analysis for different HPV groups.



```{r 5-year survival analysis}
clinical_all_CDEs <- fread("clinical_All_CDEs.txt", sep = "\t", header = TRUE)

# Transpose the clinical data
clinical_transposed <- t(clinical_all_CDEs)

clinical_transposed <- as.data.frame(clinical_transposed)

colnames(clinical_transposed) <- clinical_transposed[1,]

clinical_transposed <- clinical_transposed[-1,]

# Select necessary variables
clinical_focused_for_death <- clinical_transposed[, c(3, 4, 9, 46, 54, 114)]

# Filter data
clinical_focused_filtered_for_death <- clinical_focused_for_death %>%
  filter(hpv_status != "indeterminate", tumor_tissue_site == "head and neck")

# Convert days_to_death to numeric
clinical_focused_filtered_for_death$days_to_death <- as.numeric(clinical_focused_filtered_for_death$days_to_death)

# Define 5-year survival (1825 days)
five_years <- 1825

# Calculate 5-year survival status
clinical_focused_filtered_for_death$survived_5_years <- ifelse(clinical_focused_filtered_for_death$vital_status == "alive" | clinical_focused_filtered_for_death$days_to_death >= five_years, TRUE, FALSE)

# Calculate 5-year survival rate for HPV positive and negative groups
hpv_positive_5yr_survival_rate <- mean(clinical_focused_filtered_for_death$survived_5_years[clinical_focused_filtered_for_death$hpv_status == "positive"], na.rm = TRUE)
hpv_negative_5yr_survival_rate <- mean(clinical_focused_filtered_for_death$survived_5_years[clinical_focused_filtered_for_death$hpv_status == "negative"], na.rm = TRUE)

# Print the survival rates
cat("5-year survival rate for HPV positive group:", hpv_positive_5yr_survival_rate, "\n")
cat("5-year survival rate for HPV negative group:", hpv_negative_5yr_survival_rate, "\n")





# Calculate survival time and event status
clinical_focused_filtered_for_death$survival_time <- ifelse(
  clinical_focused_filtered_for_death$vital_status == "alive",
  five_years,
  clinical_focused_filtered_for_death$days_to_death
)

clinical_focused_filtered_for_death$event_observed <- clinical_focused_filtered_for_death$vital_status == "dead"

# Create the survival object
surv_object <- Surv(time = clinical_focused_filtered_for_death$survival_time, event = clinical_focused_filtered_for_death$event_observed)

# Fit the Kaplan-Meier curve
fit <- survfit(surv_object ~ hpv_status, data = clinical_focused_filtered_for_death)

km_plot <- ggsurvplot(
  fit,
  data = clinical_focused_filtered_for_death,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  ggtheme = theme_minimal(),
  xlab = "Days",
  ylab = "Survival Probability",
  xlim = c(0, 2500),
)

km_plot

# Save the plot
ggsave("Kaplan_Meier_Survival_Curve_Truncated.png", plot = km_plot$plot, dpi = 300, width = 10, height = 6)

```




```{r Adjust netwoek image dpi}
# Load the image
img <- image_read("network1.jpg")

img_scaled <- image_scale(img, "1024x1024")

# Save the image
image_write(img_scaled, path = "network1_scaled_300dpi.jpg", density = 300)
```





